Primosomal protein PriC rescues replication initiation stress by bypassing the DnaA-DnaB interaction step for DnaB helicase loading at oriC

Version 1

Abstract

Abstract

In Escherichia coli, replisome and replication fork assembly are initiated by DnaB helicase loading at the chromosomal origin oriC via its interactions with the DnaA initiator and the DnaC helicase loader. Upon replication fork arrest, the replisome including DnaB dissociates from the stalled fork. Replication fork progression is rescued by primosomal protein PriA- or PriC-dependent pathway in which PriA and PriC promote reloading of DnaB in different mechanisms. However, the mechanism responsible for rescue of blocked replication initiation at oriC remains unclear. Here, we found that PriC rescued blocked replication initiation in cells expressing an initiation-specific DnaC mutant, in mutant cells defective in DnaA-DnaB interactions, and in cells containing truncated oriC sequence variants. PriC rescued DnaB loading at oriC even in the absence of Rep helicase, a stimulator of the PriC-dependent replication fork restart pathway. These and results of in vitro reconstituted assays concordantly suggest that this initiation-specific rescue mechanism provides a bypass of the DnaA-DnaB interaction for DnaB loading by PriC-promoted loading of DnaB to the unwound oriC region. These findings expand understanding of mechanisms sustaining the robustness of replication initiation and specific roles for PriC in the genome maintenance.

Introduction

Chromosome replication is strictly regulated to ensure successful duplication of the genetic material (Costa & Diffley, 2022; Grimwade & Leonard, 2021; Kasho et al., 2023). In Escherichia coli, the ATP-bound form of the replication initiator protein DnaA (ATP-DnaA) forms a higher-order complex with DNA-bending protein IHF at the unique chromosomal origin oriC. This complex promotes local unwinding of oriC and recruits a pair of DnaB helicases, which are successively loaded to the single-stranded (ss) DNA regions in a bidirectional manner with the aid of DnaC helicase loader. ssDNA-loaded DnaB interacts with DnaG primase and DNA polymerase III holoenzyme, resulting in replisome (Arias-Palomo et al., 2019; Chodavarapu & Kaguni, 2016; Hayashi et al., 2020; O’Donnell et al., 2013; Sakiyama et al., 2017, 2018, 2022; Wegrzyn & Konieczny, 2023). Replisomes disassemble when the replication fork stalls in front of an obstacle, such as protein-DNA roadblock or single strand breaks. PriA recognizes the abandoned fork structure, which triggers reloading of DnaB with its partner proteins, such as PriB, PriC, DnaT, and DnaC (Heller & Marians, 2005a; Masai et al., 1994; Michel & Sandler, 2017; Windgassen et al., 2018). In addition, independently of PriA, PriC binds to the abandoned fork structure, triggering reloading of DnaB with the aid of DnaC, which is supported by Rep helicase depending on the fork structure (Heller & Marians, 2005b; Sandler et al., 1999). Also, homologous recombination mechanisms can rescue the abandoned replication fork during SOS responses (Asai & Kogomal, 1994; Kogoma, 1997; Né Dicte Michel et al., 2001). In contrast to these well characterized mechanisms at abandoned forks, little is known about what happens when replication initiation is impeded at oriC.

E. coli oriC comprises a Duplex-Unwinding Element (DUE) and a DnaA-Oligomerization Region (DOR) (Figure 1A) (Grimwade & Leonard, 2021; Kasho et al., 2023; Wegrzyn & Konieczny, 2023). The DUE contains 13-mer AT-rich sequence repeats known as L-, M-, and R-DUEs. The M- and R-DUEs are essential for stable DUE unwinding with the specific sequences (TT[A/G]T(T)) used for DnaA-ssDNA binding (Ozaki & Katayama, 2012; Sakiyama et al., 2017). The L-DUE containing the TTATT sequence promotes efficient DnaB loading by expanding the unwound oriC region (Sakiyama et al., 2022). The DOR is divided into three subregions: the Left-, Middle-, and Right-DORs, which contain asymmetric 9-mer DnaA binding sequences (DnaA box) with the consensus sequence TTA[T/A]NCACA (Figure 1A) (Noguchi et al., 2015; Ozaki & Katayama, 2012; Rozgaja et al., 2011; Sakiyama et al., 2017; Shimizu et al., 2016). The Left-DOR contains a cluster of unidirectionally aligned DnaA boxes including high-affinity DnaA box R1, and low-affinity boxes R5M, τ1−2, and I1-2, with an IHF-binding region between R1 and R5M boxes. The Right-DOR contains an oppositely oriented DnaA box cluster including high-affinity DnaA box R4 and low-affinity boxes I3 and C1-3. These DnaA box clusters form frameworks for the Left- and Right-DnaA subcomplexes, respectively. DnaA bound to the R2 box, which solely resides in the Middle-DOR, stabilizes these DnaA subcomplexes (Miller et al., 2009; Rozgaja et al., 2011; Shimizu et al., 2016). In addition, the AT-cluster (TATTAAAAAGAA) region, which connects to the L-DUE, stimulates DnaB loading in the absence of the Right-DnaA subcomplex (Sakiyama et al., 2022).

Schematic structures of oriC and DnaA and the replication initiation mechanism.

(A) Overall structure of oriC with the AT-rich cluster. oriC (245 bp) includes the Duplex-Unwinding Element (DUE, purple bar) and the DnaA-Oligomerization Region (DOR, gray bar). DUE is composed of three AT-rich 13-mer repeats termed L, M, and R. Dark purple boxes indicate the specific sequences involved in DnaA-ssDNA binding (TT[A/G]T(T)). The AT-cluster (purple dots) flanking the DUE outside oriC is a supplementary unwinding region. The DOR contains 12 DnaA boxes (filled and open arrowheads representing sites with the full consensus TTA[A/T]NCACA sequence and sites with mismatches, respectively) and an IHF-binding region (IBR, green box). The DOR is subdivided into Left, Middle, and Right subregions. (B) Domains of DnaA. Domains I-IV are shown schematically, with amino acid residue numbers shown in each bracket. H/B motifs (V211 and R245, squares) and Arginine finger (R285, triangle) are indicated. The major functions of each domain are described on the right side of the structure. (C) DUE unwinding by the ssDUE recruitment mechanism. DnaA and IHF are indicated by red and green diagrams, respectively. When ATP-DnaA oligomerizes, DUE is unwound unstably by thermal motion and torsional stress. The M/R region of the DUE-upper strand (purple line) binds to R1-DnaA and R5M-DnaA via IHF-induced DNA bending, resulting in stable DUE unwinding. In addition to M/R-DUE, L-DUE is moderately unwound and interacts with the Right-DnaA subcomplex. (D) Mechanism for DnaB loading. Each Left-DnaA and Right-DnaA subcomplex binds to a DnaB hexamer (blue ovals) complexed with DnaC (yellow circles). (E) PriC-dependent fork rescue pathways. PriC (vivid red circle) promotes DnaB reloading to the abandoned fork, which exposes ssDNA regions. PriC binds to ssDNA at the replication fork, which remodels the SSB-ssDNA complex (for simplicity, SSB is omitted) for DnaC-dependent reloading of DnaB (upper panel). If the ssDNA region is accompanied with a nascent lagging strand, Rep helicase (green shape) is recruited by PriC and unwinds the lagging strand to expose the ssDNA region for DnaB loading (the lower case).

DnaA comprises four functional domains (Figure 1B). Domain I binds to multiple proteins such as DnaB helicases and the DnaA-assembly stimulator DiaA (Abe et al., 2007; Hayashi et al., 2020; Keyamura et al., 2007, 2009). Domain II is a flexible linker (Abe et al., 2007; Nozaki & Ogawa, 2008). Domain III contains AAA+ (ATPases Associated with various cellular Activities) motifs, which are involved in tight ATP/ADP binding, ATP hydrolysis, and domain III-III interactions (Duderstadt et al., 2011; Erzberger et al., 2006; Felczak & Kaguni, 2004; Kawakami et al., 2005; Ozaki et al., 2012). The arginine finger motif (Arg285) in domain III interacts with ATP bound to domain III of adjacent DnaA protomers, stimulating DnaA complex formation in a cooperative manner (Kawakami et al., 2005; Noguchi et al., 2015). In addition, H/B motifs (Val 211 and Arg245) in this domain bind to ssDNA in a sequence-specific manner (Ozaki et al., 2008; Sakiyama et al., 2017). Domain IV, which comprises the helix-turn-helix motif, recognizes the DnaA box (Fujikawa et al., 2003).

For replication initiation, ATP-DnaA molecules cooperatively oligomerize at oriC to form the Left- and Right-DnaA subcomplexes (Figure 1C). Formation of these complexes is stimulated by homotetrameric DiaA protein, which binds up to four DnaA molecules via domain I, including Phe46 (Keyamura et al., 2007, 2009). In concert with DNA superhelicity and thermal energy, the DUE undergoes initial unstable unwinding, which is stabilized through specific interaction between H/B motifs of domain III in the Left-DnaA subcomplex and the upper strand of M- and R-DUEs. This interaction is facilitated by sharp DNA bending by IHF, resulting in stable DUE unwinding (Figure 1C) (Ozaki et al., 2008; Sakiyama et al., 2017). When the single-stranded region is expanded to the L-DUE, the resultant single- stranded L-DUE binds the H/B motifs within the Right-DnaA subcomplex, maximizing the efficiency of the DnaB loading process (Sakiyama et al., 2022).

Subsequently, two DnaB helicases are loaded onto each ssDUE strand through interactions with DnaC and DnaA (Figure 1D). DnaC binding to DnaB changes the closed ring structure of the DnaB hexamer to an open spiral form to allow it to encircle ssDNA (Arias-Palomo et al., 2019; Nagata et al., 2020). Each Left- and Right-DnaA subcomplex binds a DnaB-DnaC complex via high-affinity interaction between DnaA Phe46 and DnaB Leu160 and guides the DnaB-DnaC complexes to unwound ssDUE strands via a low-affinity interaction between DnaA domain III His136 and DnaB (Figure 1B and 1D). Upon interaction with ssDNA, DnaC dissociates from DnaB, which enables migration of DnaB hexamer with helicase action. The pair of DnaB helicases loaded onto the ssDNA strands progress in opposite directions, forming replisomes with DnaG and Pol III holoenzyme (Chodavarapu & Kaguni, 2016; O’Donnell et al., 2013).

Replication restart pathways ensure fork progression of the entire chromosome under conditions that trigger replisome disassembly (Heller & Marians, 2005a; Lopper et al., 2007; Michel & Sandler, 2017). PriA helicase-dependent pathways predominantly facilitate DnaB reloading onto abandoned forks in vivo (Flores et al., 2002). PriA is a 3’-5’ DNA helicase and has specific affinity for forked DNA structures comprising one parental dsDNA and two newly synthesized sister dsDNA strands. This permits PriA recognition of the abandoned replication fork and the subsequent unwinding of the nascent lagging strand (Duckworth et al., 2023; Windgassen et al., 2018; Windgassen & Keck, 2016). The resultant unwound ssDNA associates with PriB and DnaT to promote reloading of DnaB helicase with the aid of DnaC (Duckworth et al., 2023; Heller & Marians, 2005a; Lopper et al., 2007). In the absence of PriB, PriC participates in a PriA-dependent pathway (Sandler et al., 1999). In addition, PriA is required for oriC/DnaA-independent chromosomal replication called stable DNA replication (SDR) (Masai et al., 1994), which is promoted by UV irradiation (inducible iSDR) and loss of rnhA or recG (constitutive cSDR) (Kogoma, 1997). In both mechanisms, replication is initiated at triple-stranded structures comprising a dsDNA-ssDNA hybrid (D-loop) or a dsDNA-ssRNA hybrid (R-loop) generated at specific chromosomal loci. These structures mimic the abandoned fork structure, thereby allowing helicase loading to occur similarly to that during PriA-dependent fork restart (Masai et al., 1994).

Independently of PriA, PriC can restart replication from abandoned forks through its interaction with the ssDNA region and SSB (Single-Stranded DNA Binding protein) (Figure 1E) (Heller & Marians, 2005a; Wessel et al., 2013, 2016). PriC consists of N-terminal and C-terminal domains, which are composed of α helices and are connected by a short linker. The PriC C-terminal domain remodels the SSB-ssDNA complex, which is a prerequisite for the recruitment of DnaB helicase (Heller & Marians, 2005a; Wessel et al., 2013, 2016). PriC also interacts with Rep helicase. When the length of the ssDNA gap within the abandoned fork is short, Rep helps expand the ssDNA gap, promoting PriC-mediated remodeling of the SSB-ssDNA complex (Heller & Marians, 2005a, 2005b; Nguyen et al., 2021). Notably, in addition to these functions at the abandoned forks, PriC is suggested to play a role in DnaA-oriC- dependent replication initiation under challenging conditions, as evidenced by the synthetic lethality between priC303::kan and a subgroup of temperature-sensitive dnaA mutant alleles (dnaA46 and dnaA508): i.e. priC303::kan mutants bearing dnaA46 or dnaA508, but not wild-type (WT) dnaA, can not grow even at 30°C (Hinds & Sandler, 2004). However, the mechanisms via which PriC rescues blocked replication initiation remain elusive.

In this study, we analyzed the PriC-dependent promotion of replication initiation from oriC in the context of various replication initiation stresses. We found that PriC stimulated replication initiation in dnaC2 cells, in which DnaB helicase loading at oriC is defective, indicating that it has a role in the rescue of blocked replication initiation. Moreover, PriC stimulated the growth of cells defective in DnaA-DnaB interactions in addition to the growth of cells with oriC sequence deletions that inhibit DnaB loading. Consistent with these results, in an in vitro reconstituted system, PriC stimulated DnaB loading when the DnaA-DnaB interaction was inhibited at oriC. Furthermore, we found that PriC did not stimulate initiation of cSDR, demonstrating that PriC functions specificity in the DnaA-oriC system. Taken together, we suggest that PriC rescues blocks in replication initiation by bypassing DnaB loading by oriC-DnaA complexes.

Results

PriC stimulates DNA replication initiation in dnaA46 and dnaC2 cells

The temperature-sensitive dnaA46 and dnaA508 mutants exhibit synthetic lethality with priC303::kan even at 30°C (Hinds & Sandler, 2004), suggesting that PriC can rescue blocks in DnaA- dependent replication initiation. However, the mechanism underlying PriC-promoted rescue of blocks in replication initiation remains a mystery. Based on previous reports demonstrating PriC binding to short oligo-ssDNA, SSB, and DnaB and its role in DnaB loading at stalled replication forks (Wessel et al., 2013, 2016), we considered the possibility that PriC permits bypass of DnaA-dependent stable DUE unwinding and/or DnaB loading at the oriC ssDUE. To genetically test these ideas, we analyzed the impact of PriC on the growth of a temperature-sensitive dnaC2 mutant, which is competent for ongoing replisome progression at 37–42°C, but defective in DnaB loading specifically at the oriC ssDUE (Withers & Bernander, 1998).

Colony formation of dnaA46 cells was moderately inhibited at 37℃ (Figure 2A). Consistent with a previous report, when crossed with the ΔpriC allele, colony formation of dnaA46 mutant cells but not WT dnaA cells was severely inhibited at 37℃ (Figure 2A). Unlike the dnaA46 priC303::kan mutant, dnaA46 ΔpriC cells formed colonies at 30°C, although colony growth was slightly retarded.

Requirement of priC for DNA replication initiation in dnaA46 and dnaC2 strains.

(A) Colony forming abilities. MG1655 and its derivatives MIT125 (MG1655 dnaA46 tnaA::Tn10), KYA018 (MG1655 dnaC2 zjj18::cat), KRC002 (MG1655 ΔpriC::frt-kan), KRC004 (MG1655 dnaA46 tnaA::Tn10 ΔpriC::frt-kan), and KRC005 (MG1655 dnaC2 zjj18::cat ΔpriC::frt-kan) were grown overnight. 10-fold-serial dilutions of the overnight cultures (∼109 cells/mL) were incubated for 16 h at 30℃, or for 14 h at 35/37/42℃ on LB agar medium. Three independent experiments were performed. +, WT priC; -, ΔpriC::frt-kan. The images of representative plates are shown.

(B) DNA replication initiation in synchronized dnaA46 cells with or without WT priC. Exponentially growing MIT125 (MG1655 dnaA46 tnaA::Tn10) (+) and KRC004 (MG1655 dnaA46 tnaA::Tn10 ΔpriC::frt-kan) (−) cells were synchronized to the pre-initiation stage by incubation for 90 min at 42℃ in LB medium, followed by incubation for 5 min at 30℃ to initiate DNA replication, which was further followed by incubation with rifampicin and cephalexin for 4 h at 42℃ to allow run-out of ongoing DNA replication. Samples were withdrawn after synchronization and run-out replication. The DNA contents of these samples were measured by flow cytometry. Representative histograms of three independent experiments are shown. For each sample, the mean number of cells with more than one chromosome equivalent after a 5 min release was quantified and is shown as a percentage with standard deviations.

(C) DNA replication initiation in synchronized dnaC2 cells with or without WT priC. Exponentially growing KYA018 (MG1655 dnaC2 zjj18::cat) (+) and KRC005 (MG1655 dnaC2 zjj18::ca ΔpriC::frt- kan) (−) cells were synchronized to the pre-initiation stage by incubation for 80 min at 37℃ in LB medium, followed by incubation for 5 min at 30℃ and further incubation with rifampicin and cephalexin as described above. Samples were withdrawn and analyzed as described above.

This difference may stem from the different strain backgrounds. Notably, colony formation of dnaC2 cells at 35℃ was moderately inhibited, whereas it was severely inhibited in dnaC2 ΔpriC cells (Figure 2A). At 30°C, colony formation by dnaC2 cells and dnaC2 ΔpriC cells was similar. The observed requirement for PriC for colony formation by dnaC2 cells at 35℃ supports the idea that PriC assists in the DnaC-dependent DnaB loading step at oriC.

To investigate replication initiation from oriC, we used flow cytometry of synchronized cells bearing dnaA46 or dnaC2 with or without ΔpriC. In these experiments, cells were incubated at 42°C or 37°C for 80–90 min to synchronize the replication cycle at the pre-initiation step, after which incubation was continued at 30°C for 5 min to allow initiation and then 4 h in the presence of rifampicin and cephalexin to inhibit both replication initiation and cell division and allow replication run-out of the whole chromosome. The resultant number of chromosomes per cell, measured by flow cytometry, is known to correspond to the number of oriC copies per cell at the time of drug addition, an indicator of initiation activity.

A majority of the synchronized dnaA46 and dnaC2 cells had only a single oriC copy per cell after incubation at 42°C or 37°C regardless of the presence of ΔpriC (Figure 2B and 2C). In WT priC cells, after release of initiation at 30°C for 5 min, most cells had two oriC copies. However, in dnaA46 ΔpriC cells, only about half of the cells had two oriC copies after the release (Figure 2B), indicating moderate inhibition of initiation. In dnaC2 ΔpriC cells, the number of two-oriC cells was slightly lower than that of dnaC2 cells (Figure 2C). These observations are consistent with the colony formation data and with our hypothesis that PriC contributes to the rescue of blocked replication initiation by assisting DnaB loading onto the oriC ssDNA.

PriC rescues the cell growth of DnaA mutants defective in DnaB interactions

DnaA contains two regions for interactions with DnaB during DnaB loading at oriC ssDNA: one is a high-affinity region in domain I containing Phe46, and the other is a low-affinity region in domain III containing His136. To analyze the mechanism of PriC-promoted DnaB loading at oriC, we introduced a series of pING1 vector-based plasmids encoding WT DnaA (pKA234) or the DnaA F46A H136A double mutant (pFH) into dnaA46 cells with or without WT priC. Growth of dnaA46 cells containing pING1 was inhibited at 42℃, but the leaky expression of dnaA from pKA234 enabled the growth of these cells, as previously reported (Sakiyama et al., 2018) (Figure 3). Introduction of pFH moderately supported colony formation by dnaA46 cells at 42℃. By contrast, dnaA46 ΔpriC cells carrying pFH exhibited little growth at 42℃ (Figure 3). These results suggest that PriC is crucial for the growth of cells defective in specific DnaA-DnaB interactions, possibly because it can bypass the DnaA- DnaB interaction step required for DnaB loading onto oriC ssDNA.

Role of PriC in growth of DnaA mutant defective in interactions with DnaB.

Colony forming abilities of MIT125 (MG1655 dnaA46 tnaA::Tn10) and KRC004 (MG1655 dnaA46 tnaA::Tn10 ΔpriC::frt-kan) cells bearing pING1 plasmid vector (−) or its derivatives pKA234 expressing WT DnaA (WT) or pFH expressing DnaA F46A H136A (F46A H136A). Cells were grown at 30°C overnight and 10-fold-serial dilutions of the cultures (∼109 cells/mL) were incubated on LB agar plates containing ampicillin for 16 h at 30°C and for 14 h at 42°C, respectively. Four independent experiments were performed. +, WT priC; -, ΔpriC::frt-kan.

PriC rescues DNA replication inhibition by DiaA overexpression in vivo

DiaA, a homotetrameric protein, binds to multiple DnaA molecules and stimulates DnaA multimer assembly at oriC (Keyamura et al., 2007). The Phe46 residue in DnaA domain I is part of the DiaA-binding site and DnaB binding site. Oversupply of DiaA inhibits timely initiation of replication in cells growing in tryptone medium at 30°C, probably because of binding competition between DiaA and DnaB for DnaA domain I (Ishida et al., 2004; Keyamura et al., 2009). If PriC can bypass the DnaA-DnaB interaction requirement for DnaB loading onto oriC, PriC should suppress inhibition of initiation by DiaA oversupply. To test this possibility, we introduced pBR322 or its derivative encoding diaA (pNA135) into cells with or without WT priC (Figure 4). WT priC cells carrying pBR322 or pNA135 similarly formed colonies on LB agar plates containing ampicillin between 25°C and 42°C (Figure 4A and Figure 4-figure supplement 1). However, in ΔpriC cells carrying pNA135, but not pBR322, colony formation was moderately inhibited at 25°C (Figure 4A), supporting our conjecture that PriC can bypass the requirement for the DnaA-DnaB interaction.

PriC stimulation of DNA replication initiation inhibited by DiaA oversupply.

(A) Colony forming abilities of MG1655, KRC002 (MG1655 ΔpriC::frt-kan), and FF001 (MG1655 Δrep::frt-kan) cells bearing pBR322 or pNA135 (pBR322 derivatives carrying diaA). Overnight cultures (∼109 cells/mL) of MG1655 and KRC002 cells bearing pBR322 or pNA135 were 10-fold serially diluted and incubated on LB agar plates containing ampicillin for 40 h at 25℃. The results of three independent experiments were consistent, and one which is shown.

(B) Flow cytometry analyses of MG1655 and KRC002 (MG1655 ΔpriC::frt-kan) cells bearing pBR322 or pNA135. Cells were exponentially grown at 30℃ in LB medium with ampicillin, followed by further incubation with rifampicin and cephalexin to allow run-out of chromosomal DNA replication. DNA contents were quantified by flow cytometry, and cell size (mass) at the time of drug addition was measured by a Coulter counter. Mean mass, ori/mass ratio, doubling time (Td), and asynchrony index (A.I.: the percentage of cell numbers with non-2n copies of oriC per the cell numbers with 2n copies of oriC) of each strain are indicated at the top right of each panel with standard deviations. Three to four independent experiments were performed. +, WT priC; −, ΔpriC::frt-kan.

(C) Flow cytometry analyses of MG1655, KRC003 (MG1655 ΔpriC::frt), SA103 (MG1655 ΔdiaA::frt-kan), and KRC006 (MG1655 ΔpriC::frt, ΔdiaA::frt-kan) cells. Cells were grown and analyzed as described above. Three independent experiments were performed. For “priC” column, +, WT priC; −, ΔpriC::frt. For “diaA” column, +, WT diaA; −, ΔdiaA::frt-kan.

Colony formation of cells carrying pBR322 or pNA135 with WT priC or ΔpriC::frt-kan.

Colony forming abilities of MG1655 or KRC002 (MG1655 ΔpriC::frt-kan) cells bearing pBR322 or pNA135 (pBR322 derivatives carrying diaA). Overnight cultures (∼109 cells/mL) of MG1655 and KRC002 cells carrying pBR322 or pNA135 were 10-fold serially diluted and incubated on LB agar plates containing ampicillin for 14 h at 30–42℃. +, WT priC; −, ΔpriC::frt-kan. The results of three independent experiments were consistent, and one representative result is shown.

Rep helicase stimulates PriC-mediated rescue of arrested replication forks containing nascent strands (Heller & Marians, 2005b, 2005a; Sandler, 2000). To test the contribution of Rep helicase to PriC-promoted rescue of blocked replication initiation, we introduced pBR322 or pNA135 into Δrep cells. Unlike ΔpriC cells, Δrep cells carrying pNA135 grew similarly to Δrep cells carrying pBR322 (Figure 4A), suggesting no involvement of Rep in the PriC-promoted rescue of blocked replication initiation. This is consistent with the features of unwound oriC, which has a fork structure similar to that of the replication fork but without nascent strands, and the specific role of Rep in rescuing abandoned replication forks with nascent strands in the PriC-dependent pathway. Rep is specifically required for the unwinding of the nascent leading strand of abandoned replication forks, which is needed to expand the ssDNA region required for entry of PriC (Heller & Marians, 2005b, 2005a).

We further analyzed replication initiation in ΔpriC cells carrying pNA135 at 30°C by flow cytometry (Figure 4B). Exponentially growing cells were further incubated for run-out replication after the addition of rifampicin and cefalexin. The volume (mass) of cells at the time of drug addition was also quantified. Under these conditions, WT priC cells bearing pBR322 mainly had eight or four oriC copies per cell with eight-oriC cells predominating (Figure 4B). The number of oriC copies per cell was similar to that in WT priC cells bearing pNA135, with only slight increase of asynchronous initiation, as shown by the slight increase in the number of cells with non-2n oriC copies, which is basically consistent with our previous data (Ishida et al., 2004). However, in ΔpriC cells bearing pNA135, severe inhibition of initiation occurred, as shown by the substantial decrease in the number of eight-oriC cells and the large increase of the number of cells with only one to three oriC copies (Figure 4B). This result supports the proposed function of PriC at oriC mentioned above. In addition, the observed asynchronous initiations in ΔpriC cells bearing pBR322 is worthy of attention. Indeed, the fraction of cells with non-2n oriC copies was higher among ΔpriC cells than among WT priC cells (Figure 4B), suggesting that even in normal cells, inhibition of replication initiation occurs at low frequency and requires PriC function to rescue replication initiation.

To test whether PriC rescues replication initiation inhibited by deletion of the diaA gene, we conducted similar flow cytometry analysis of ΔpriC, ΔdiaA, and ΔpriC ΔdiaA cells (Figure 4C). ΔdiaA cells showed severe asynchrony of replication initiation, as shown by the increase in the number of five- to-seven-oriC cells, which supports the stimulatory role of DiaA in DnaA-oriC complex formation (Figure 4C). However, ΔpriC ΔdiaA cells did not show more asynchronization of replication initiation than ΔdiaA cells, as shown by the similar asynchrony index (A.I.). These results suggest that PriC does not support or bypass DnaA-oriC complex formation.

PriC stimulates replication initiation in cells with oriC sequence truncations

Right DnaA-subcomplex plays a stimulatory role in DnaB helicase loading (Sakiyama et al., 2022). The loading of two DnaB hexamers at oriC ssDNA regions requires multiple steps that depend on the distinct functions of two DnaA subcomplexes (Figure 1): the Left-DnaA subcomplex stably unwinds M-R regions of the DUE and the Right-DnaA subcomplex expands the unwound region to the AT-L region, supporting efficient DnaB helicase loading (Sakiyama et al., 2017, 2022; Yoshida et al., 2023). The expanded ssDNA regions in this open oriC complex efficiently promote the loading of one DnaB hexamer onto the lower (A-rich) strand M-R region and a second DnaB hexamer onto the upper (T-rich) strand of the DUE (Fang et al., 1999; Sakiyama et al., 2022). In the absence of the Right-DnaA subcomplex, the AT region of the AT-L region assists in DnaB loading (Sakiyama et al., 2022; Stepankiw et al., 2009)

To determine whether PriC rescues defects in DnaB loading when the oriC sequence is altered to prevent Right-DnaA subcomplex formation, we first assessed the colony formation of cells lacking the Middle- and Right-DOR of oriC (Left-oriC) (Figure 5A). The Left-oriC mutant formed colonies at 37°C but showed severe growth defects at 25°C and 30°C (Figure 5B). At 37°C, thermal energy might be sufficient to expand DUE unwinding into the AT region and allow DnaB loading in the absence of the Right-DnaA subcomplex (Sakiyama et al., 2022), but not at 25°C or 30°C. By contrast, colony formation of the Left-oriC ΔpriC double mutant was markedly compromised at 37°C (Figure 5B) and the growth defects of the Left-oriC mutant 25°C and 30°C were aggravated. These findings support the idea that PriC functions to rescue defective processes in DnaB loading caused by deletions of the oriC sequence that prevent DnaA subcomplex formation.

Stimulation of replication initiation by PriC in Right-DnaA subcomplex-defective cells

(A) Schematic structure of Left-oriC and R4Tma oriC. AT-rich region, DUE, and DOR are indicated by a gray dotted box, a gray box, and an open box, respectively. Filled and open arrowheads in the DOR show DnaA boxes with the full consensus sequence or mismatches. In addition, a striped box indicates the IBR. The regions of each mutant oriC are indicated below the structure as open boxes. The position of the TmaDnaA box substitution is indicated by a yellow arrowhead.

(B) Colony forming abilities of WT oriC and Left-oriC cells with or without WT priC. Overnight cultures (∼109 cells/mL) of NY20-frt (WT oriC; WT) and NY20L-frt (Left-oriC; Left) cells with (+) or without (−) WT priC were 10-fold serially diluted and incubated on LB agar plates for 24 h at 25°C, for 16 h at 30°C, and for 14 h at 37°C. +, WT priC; −, ΔpriC::frt-kan. Three independent experiments were performed.

(C) Colony forming abilities of WT oriC and oriC R4-box mutant cells with or without WT priC. NY20-frt (WT oriC) and NY24-frt (oriC with R4Tma substitution, R4Tma) cells with (+) or without (−) WT priC were grown as described in panel A. +, WT priC; −, ΔpriC::frt-kan. Four independent experiments were performed.

(D) Flow cytometry analyses of WT and R4Tma cells with or without WT priC. Cells were exponentially grown at 30℃ in LB medium with ampicillin and analyzed as described in Figure 4. Representative histograms from four independent experiments are shown. Mean mass, ori/mass, and Td of each strain are indicated at the top right of each panel with standard deviations.

Next, similar experiments were performed using a mutant in which the R4 DnaA box within Right-DOR was substituted with a sequence (R4Tma) defective in binding of E. coli DnaA (Figure 5A and 5C) (Noguchi et al., 2015; Sakiyama et al., 2022). We reasoned that formation of the Right-DnaA subcomplex would be inhibited by the introduction of R4Tma and PriC would be required for robust initiation in this context. As in the Left-oriC strain, the R4Tma strain showed normal colony formation in the presence of WT priC. By contrast, in the R4Tma ΔpriC strain, colony formation was impaired moderately at 25°C and slightly at 30°C (Figure 5C), indicating that PriC can rescue the blocked replication initiation caused by the absence of the R4 box. The reduced colony formation of R4Tma ΔpriC double mutant cells was alleviated at higher temperatures, similar to the phenotype of the Left-oriC mutant.

Furthermore, we assessed replication initiation in R4Tma cells at 30°C using flow cytometry (Figure 5D). R4Tma cells grew slightly slower than WT oriC cells and showed clear inhibition of initiation; in contrast to the WT oriC strain, the R4Tma mutant cell population contained more four-oriC cells and fewer eight-oriC cells, and showed severe asynchronous initiation. Notably, these negative effects of the R4Tma mutation were amplified by deletion of priC, i.e., the number of one to three-oriC cells increased and the ori/mass ratio was further reduced.

We next analyzed the role of PriC in cells lacking the AT-L region of oriC (subATL oriC) (Figure 6A). L-DUE and the flanking AT region assist in DnaB helicase loading by stimulating oriC unwinding, which is essential in a strain defective in Right-DnaA subcomplex formation (Sakiyama et al., 2022). SubATL oriC cells and WT oriC cells formed colonies with similar efficiency on LB agar plates between 25°C and 37°C, regardless of the presence of ΔpriC (Figure 6B). However, flow cytometry analysis revealed that deletion of priC specifically inhibited initiation in subATL oriC cells (Figure 6C); the number of eight-oriC cells was lower while the number of four-to-seven oriC cells was higher in subATL oriC ΔpriC cells than in subATL oriC cells with WT priC. Taken together, these results are consistent with the idea that PriC restores initiation and DnaB loading at such truncated oriC sequences.

Stimulation of replication initiation by PriC in cells with a deletion in the ssDNA-expanding region of oriC

(A) Schematic structure of subATL oriC as shown in Figure 5. The oriC region including the subATL oriC is indicated below the structure as a white box.

(B) Colony forming abilities of WT oriC and oriC subATL mutant cells with or without WT priC. NY20-frt (WT oriC) and NY20ATL-frt (oriC lacking AT and L sequences, subATL) with (+) or without (−) WT priC were grown as described in Figure 5. +, WT priC; −, ΔpriC::frt-kan. Three independent experiments were performed.

(C) Flow cytometry analyses of WT oriC and oriC subATL mutant cells with or without WT priC. Cells were exponentially grown at 30℃ in LB medium and analyzed as described in Figure 4. Representative histograms from three independent experiments are shown. Mean mass, ori/mass, and Td of each strain are indicated at the top right of each panel with standard deviations.

PriC loads DnaB at oriC unwound by the DnaA complex in an in vitro reconstituted system

To analyze the mechanism of PriC rescue of blocked initiation at oriC, we assessed PriC activity in an in vitro reconstituted system for DnaB loading and DnaB-promoted DNA unwinding using the supercoiled circular form (form I) of the oriC plasmid pBSoriC and purified proteins, namely DnaA, N- terminally histidine-tagged DnaB (His-DnaB), DnaC, IHF, SSB, and gyrase. In this system, DnaA and IHF unwind the DUE and stimulate His-DnaB loading to the ssDUE region with the aid of DnaC. The loaded His-DnaB expands the ssDNA regions in concert with the ssDNA-binding activities of SSB. Concomitantly, the positive supercoiling generated through DNA unwinding by His-DnaB is resolved by DNA gyrase, resulting in the formation of plasmid DNA topoisomers (form I*). Form I* can be separated from form I by agarose gel electrophoresis (Sakiyama et al., 2018, 2022).

First, we assessed the effect of PriC on WT DnaA. WT DnaA stimulated form I* formation in the absence of PriC (Figure 7A, lanes 1 and 5) and increasing concentrations of PriC moderately inhibited WT DnaA-mediated form I* formation (Figure 7A, lanes 5 to 8), probably because of competition between DnaA and PriC for binding to DnaB. Next, we focused on the DnaA F46A H136A double mutant. The DnaA F46A and DnaA H136A mutant proteins alone are reported not to support form I* formation (Keyamura et al., 2009; Sakiyama et al., 2018). Consistently, the DnaA F46A H136A double mutant protein was inactive in form I* formation (Figure 7A, lane 9). However, when PriC was added to the assay, DnaA F46A H136A promoted form I* formation to a level comparable to that of WT DnaA in the presence of PriC (Figure 7A, lanes 5-12 and 7B), indicating that PriC rescues defective DnaB loading by DnaA F46A H136A-oriC complexes. These findings are consistent with the idea that PriC can bypass the strict reliance on DnaA-DnaB interactions for DnaB loading at oriC.

DnaB loading onto the ssDUE by PriC.

(A and B) In vitro reconstituted system for DnaB loading. Form I of oriC plasmid (pBSoriC; 1.6 nM) was incubated for 15 min at 30℃ with the indicated amount of PriC in the presence of ATP-DnaA or its mutant derivatives, DnaB, DnaC, IHF, gyrase, and SSB. The resulting plasmids were purified and analyzed using agarose gel electrophoresis. A representative image of three independent experiments is shown in the black/white-inverted mode (A). Band intensities of each lane in the gel image were analyzed by densitometric scanning. The percentages of form I* oriC plasmid per input DNA are shown as ‘Form I* (%)’. Mean and standard deviations are shown (B). Abbreviations in panel B: −, no addition of DnaA; WT, wild-type DnaA; FH, DnaA F46A H136A; VA, DnaA V211A.

(C and D) In vitro reconstituted system containing DiaA. Form I of pBSoriC plasmid was used as described above except for the addition of the indicated amounts of DiaA. A representative image of three independent experiments is shown in black/white-inverted mode (C). Band intensities of each lane in the gel image were analyzed by densitometric scanning. The percentages of form I*oriC plasmid together with the mean and standard deviations are shown as described above (D).

DUE unwinding activities of DnaA mutant deriviatives in the presence and absence of PriC.

(A and B) in vitro DUE unwinding assays using DnaA mutant derivatives. Form I of oriC (pBSoriC; 1.6 nM) was incubated for 9 min at 30℃ with the indicated amount of DnaA or its mutant derivatives in the presence of IHF (42 nM), followed by incubation with P1 nuclease (1.5 units) and further incubation with AlwNI before agarose gel electrophoresis. Note that 2.6 kb and 1.0 kb fragments are observed with restriction enzyme digestion when the ssDNA-specific P1 nuclease cleaves 3.6 kb pBSoriC at the DUE. A representative image of two independent experiments is shown in black/white-inverted mode (A). The sizes of each band are indicated at the left side of the image. Bands intensities of each lane in the gel image were analyzed by densitometric scanning. The percentages of P1 nuclease-digested oriC DNA per input DNA molecules are shown as ‘DUE unwinding (%)’. Mean and data are shown (B). Abbreviations in panel B: −, no addition of DnaA; WT, wild-type DnaA; FH, DnaA F46A H136A; VA, DnaA V211A. (C and D) In vitro DUE unwinding assay with PriC. pBSoriC was used as described above except for PriC addition. A representative image of two independent experiments is shown in black/white-inverted mode (C). Data were analyzed as described above (D). n.t., not tested.

To corroborate this idea, we also determined whether PriC rescues blocked initiation resulting from unstable DUE unwinding. oriC complexes containing DnaA V211A cannot bind to the ssDUE and consequently fails in stable DUE unwinding (Ozaki et al., 2008). The DnaA V211A mutant was virtually inactive for form I * formation irrespective of the presence or absence of PriC (Figure 7A, lanes 13-16, and 7B), indicating that PriC cannot efficiently rescue a defect of DUE unwinding.

We further evaluated the DnaB loading activity of PriC in the presence of DiaA using a similar in vitro reconstituted system. Previously, DiaA was shown to stimulate oriC DUE unwinding, but to inhibit form I* formation, because DiaA-DnaA binding competitively inhibits DnaB-DnaA binding; DiaA and DnaB share the same binding site in DnaA domain I (Figures 7C lanes 1, 3, 5, and 7 and 7D) (Keyamura et al., 2009). PriC moderately stimulated form I* formation specifically in the presence of both DnaA and DiaA (Figures 7C lanes 7 and 8, and 7D). These results further support the idea that PriC can bypass the specific requirement for the DnaA-DnaB interaction in DnaB loading at oriC. DiaA bound to oriC-DnaA complexes might be a physical obstacle reducing efficiency of DnaB loading by PriC.

In addition, we analyzed the DUE unwinding activities of DnaA mutants using the in vitro reconstituted system. As previously reported, DnaA V211A is largely inactive in DUE unwinding, but DnaA F46A H136A unwinds the DUE at a level comparable to that of WT DnaA (Figure 7-figure supplement 2A and 2B). PriC did not stimulate DUE unwinding irrespective of the presence of WT DnaA (Figure 7-figure supplement 2C and 2D). These results demonstrate the functional specificity of PriC in rescuing failed DnaB loading at oriC ssDNA and are consistent with the PriC mechanism mentioned above.

Role for PriC in cSDR

Finally, to analyze the role of PriC in other types of replication initiation, we examine whether PriC contributes to the initiation of cSDR, which does not require dnaA or oriC and is activated in cells lacking rnhA or recG. Although the PriA-PriB-DnaT primosome complex has previously been reported to be involved in the initiation of cSDR (Heller & Marians, 2006), it is unclear whether PriC has any role in this type of initiation.

Unlike dnaA46 cells, dnaA46 rnhA::cat double mutant cells formed colonies even at 40°C (Figure 8A), indicating that these cells were engaged in cSDR, as previously reported (Hinds & Sandler, 2004). The formation of colonies by ΔpriC dnaA46 rnhA::cat triple mutant cells was severely inhibited at 40°C (Figure 8A), suggesting that PriC contributes to the growth of dnaA46 rnhA::cat mutant cells.

PriC stimulation of cSDR-dependent cell growth.

(A) Colony forming abilities. Cells of MIT125 (MG1655 dnaA46), MIT125c (MG1655 dnaA46 rnhA::cat), KRC004 (MG1655 dnaA46 ΔpriC::frt-kan), and KRC004c (MG1655 dnaA46 ΔpriC::frt-kan rnhA::cat) were grown at 30℃ overnight and 10-fold-serial dilutions of the overnight cultures (∼109 cells/mL) were incubated for 16 h at 30℃ and for 36 h at 40℃ on LB agar plates. Three independent experiments were performed. +, WT; −, deletion.

(B) oriC and ter copy numbers. MIT125, MIT125c, KRC004, and KRC004c cells growing exponentially at 30℃ were further incubated for 90 min at 40℃ in LB medium. Samples were withdrawn before and after the 40℃ incubation. The genome DNA of each sample was extracted by boiling the cells for 5 min at 95℃. The relative copy numbers of oriC (84.6 min) and ter (32.4 min) to that of yapB (50.5 min) were quantified using real-time qPCR. The data and averages of three independent experiments are shown with standard deviations.

To determine whether PriC contributes to the initiation of dnaA46-oriC replication or cSDR, we calculated the relative ratios of oriC (84.6 min) and ter (32.4 min) copy numbers to the refence locus yapB (50.5 min) copy number using real-time quantitative PCR (qPCR). To calculate the copy number of the chromosomal ter region, the preferential initiation site of cSDR (Maduike et al., 2014), dnaA46 rnhA::cat cells growing exponentially at 30°C were shifted to 40°C for 90 min and the copy number before and after incubation at 40°C was calculated. Consistent with the colony formation data (Figure 8A), the ter copy number ratio of the dnaA46 rnhA::cat double mutant increased after 40°C incubation, whereas the copy number ratio of the dnaA46 mutant expressing WT rnhA remained the same. The ter copy number ratio of the dnaA46 rnhA::cat ΔpriC triple mutant strain was also increased, indicating that cSDR could occur even in the absence of PriC.

To monitor initiation of DnaA-oriC replication, we determined the oriC copy number ratio (Figure 8B). At the permissive temperature, the oriC copy number ratio of the dnaA46 ΔpriC double mutant strain was lower than that of the dnaA46 single mutant, confirming the importance of PriC for replication initiation in dnaA46 cells, as shown in Figure 2. In the dnaA46 rnhA::cat double mutant expressing WT priC, the oriC copy number ratio was reduced, which is explained by activation of cSDR and increase of the relative copy number of the reference yapB locus (50.5 min chromosomal map position). Together, these results suggest the idea that PriC stimulates replication fork progression during cSDR.

Discussion

In growing cells, replication initiation can be blocked under challenging conditions. However, the mechanisms that rescue blocked initiations are largely unknown. To elucidate such mechanisms, we investigated PriC-promoted replication initiation under various challenging conditions. We showed that PriC was necessary for the optimal growth of dnaC2 cells at the semi-restrictive temperature and for replication initiation even at the permissive temperature (Figure 2), suggesting that PriC contributes to replication initiation rescue in dnaC2 cells by facilitating DnaB helicase loading at oriC. Furthermore, we observed that PriC was necessary for the optimal growth of cells in which the DnaA-DnaB interaction was inhibited by DiaA overexpression or DnaA F46A H136A double mutations (Figures 3 and 4). In addition, PriC was stimulatory for the colony formation and replication initiation bearing oriC mutations that impair efficient DnaB loading (Figures 5 and 6). These results, together with results of form I* formation assays (Figures 7), suggest that PriC bypasses the need for the DnaA-DnaB interaction by binding to unwound oriC DNA and recruiting DnaB to facilitate its loading onto ssDNA (Figure 9). This mechanism is similar to that proposed for the rescue of abandoned replication forks by PriC (Figure 1E).

Model for PriC-promoted replication initiation.

PriC-promoted rescue of blocked DnaA-oriC dependent replication initiation. oriC, DnaA, IHF, DnaB, and DnaC are illustrated as depicted in Figure 1C. When tight DnaA binding of DiaA (dark gray circle) (or DnaA F46A H136A double mutations) inhibits DnaB recruitment, PriC (Brilliant red circle) binds to a stably unwound DUE strand, recruiting a DnaB-DnaC complex to lower strand of DUE instead of Left-DnaA subcomplex. When a sufficient ssDNA region is exposed, PriC then recruits the DnaB-DnaC complex to the opposite strand.

The finding that PriC failed to stimulate form I* formation in the presence of DnaA V211A, which lacks DUE unwinding activity (Figures 7), provides mechanistic insight into PriC-dependent replication initiation at oriC. We propose that PriC-dependent replication initiation possibly begins with the recognition of a stably unwound ssDNA region by PriC (Figure 9). Consistent with this perspective, PriC ssDNA-binding activity and ssDNA regions have been reported to be important for PriC-promoted replication fork rescue (Heller & Marians, 2006; McMillan & Keck, 2024). The failure of PriC to stimulate replication initiation in ΔdiaA cells (Figure 4C) is also potentially explained by the reduced DUE unwinding caused by destabilization of the DnaA-oriC complex in ΔdiaA cells. Based on this, we suggest that the stably unwound ssDNA region in oriC serves as a distinctive feature for PriC- promoting DnaB loading at oriC for replication initiation rescue, and distinguishes it from PriA- dependent replication restart, which requires the structure of the replication fork DNA consisting of sister dsDNA strands (Heller & Marians, 2005a; Windgassen et al., 2018).

In this study, we found that cells with DnaA F46A H136A double mutations grew well, but only when the cells expressed WT PriC, whereas the growth of cells expressing the DnaA H136A single mutant was severely inhibited even when the cells expressed WT PriC (Sakiyama et al., 2018). The His136 residue is located within the weak, secondary DnaB interaction region in DnaA, and is crucial for DnaB loading onto oriC ssDNA. Although domain I in DnaA H136A can stably tether DnaB-DnaC complexes to DnaA complexes on oriC (Sakiyama et al., 2018), the complexes fail to load DnaB onto oriC ssDNA even in the presence of PriC. It is possible that the interaction between PriC and DnaB is inhibited by DnaB binding to DnaA domain I. Conversely, the PriC-DnaB interaction may inhibit the interaction between DnaA domain I and DnaB, as suggested by the inhibitory effect of PriC on DnaB loading in vitro in the presence of WT DnaA (Figure 7). However, this inhibitory effect of PriC was not observed in vivo, suggesting that PriC-DnaB binding could involve still unknown factors and regulatory mechanisms.

PriC supported cSDR-dependent growth in dnaA46 rnhA::cat double mutant cells without stimulating the initiation of cSDR (Figure 8). In this mutant, cSDR predominantly initiates from the ter region, resulting in head-on collisions between the replisome and transcription complexes in rrn operons (Maduike et al., 2014). This suggests that in cSDR cells PriC specifically rescues abandoned forks including those generated during head-on conflicts between the replisome and transcription complexes.

Bacillus subtilis (B. subtilis), which expresses a PriA homolog but not a PriC homolog, employs a PriA-independent fork rescue pathway (Bruand et al., 2001). In this pathway, B. subtilis DnaC (BsuDnaC) helicase is reloaded using the helicase loader DnaI together with B. subtilis DnaB (BsuDnaB) and DnaD coloaders. The ssDNA-binding activity of BsuDnaB has a crucial role in this PriA-independent pathway (Bruand et al., 2005). SsDNA-bound BsuDnaB remodels the SSB-ssDNA complex with DnaD and recruits BsuDnaC. These functions of BsuDnaB are similar to the proposed functions of PriC in replication fork rescue in E. coli. In addition, BsuDnaC helicase is recruited to oriC by concerted actions of B. subtilis DnaA and the DnaD-BsuDnaB complex, which has an essential role in chromosome replication initiation (Jameson & Wilkinson, 2017), suggesting a similar role of BsuDnaB to PriC in E. coli cells expressing DnaA F46A H136A. These similarities between PriC and BsuDnaB in these two evolutionarily distant bacterial species suggest that the mechanism of PriC- promoted helicase loading is conserved among bacterial species despite the absence of sequence conservation of PriC and BsuDnaB homologs.

Based on our results, we propose the abnormal competition between DnaB and DiaA for DnaA domain I could represent a form of intrinsic replication initiation stress in bacteria with conserved DiaA homologs. This type of stress could also occur in ε-proteobacterial species, such as Helicobacter pylori, because they express HobA, a DiaA-functional homolog (Natrajan et al., 2007; Zawilak-Pawlik et al., 2011). Similarly, YfdR, encoded by E. coli prophages, inhibits replication initiation by competing with DnaB for binding to DnaA domain I (Noguchi & Katayama, 2016), suggesting that such extraneously introduced inhibitors could trigger replication initiation stress.

Even in human, ORC1, ORC2, and ORC5, the essential components of the eukaryotic replication initiation complex, are not essential in some cancer cell-lines (Shibata & Dutta, 2020; Struhl et al., 2016), suggesting that mechanisms of replication initiation rescue may also operate beyond the bacterial kingdom. Therefore, further investigation of the initiation rescue processes and the factors involved in diverse organisms from bacteria to human will be important for a full understanding of the common principles and diverse mechanisms that ensure robust initiation of chromosomal DNA replication.

Materials and methods

Plasmids, proteins, and strains

The plasmids used in this study are listed in Key Resource Table. pKA234, pKW44-1, pNA135, pBSoriC, pET22b(+)-priC, and pTKM601 were described previously (Aramaki et al., 2015; Ishida et al., 2004; Kawakami et al., 2005; Keyamura et al., 2009; Kubota et al., 1997; Ozaki et al., 2008). For the construction of pFH, an alanine substitution was introduced into pH136A with specific mutagenic primer sets using the QuikChange site-directed mutagenesis protocol (Stratagene [Agilent], Agilent, La Jolla, CA, United States) as described previously (Keyamura et al., 2007; Sakiyama et al., 2018). WT DnaA and its derivative proteins were overproduced in E. coli strain KA450 from pKA234, pKW44-1, or pFH and purified as described previously (Noguchi et al., 2015; Ozaki et al., 2008; Sakiyama et al., 2017).

PriC protein was prepared as previously reported with minor modifications (Aramaki et al., 2013). Briefly, PriC protein was overproduced in E. coli strain BL21-codonPlus-RIL by inducing its expression from pET22b(+)-priC using 0.2 mM isopropyl-β-D-1-thiogalactopyranoside (IPTG). The resulting cells were suspended in chilled buffer A (50 mM HEPES-NaOH [pH7.0], 10% sucrose, 1 mM EDTA, 2 mM dithiothreitol (DTT), and 1 mM PMSF) and disrupted by sonication. Proteins in the soluble fraction were precipitated with 0.24 g/mL ammonium sulfate, resuspended in a separation buffer (50 mM imidazole [pH 7.0], 20% glycerol, 2 mM DTT, 1 mM EDTA, and 40 mM ammonium sulfate) and loaded onto a 1 mL HiTrap® SP HP column (Cytiva, Uppsala, Sweden). Bound proteins were eluted with a linear gradient from 0 to 1 M sodium chloride.

All E. coli strains used in this study are listed in Key Resource Table. KYA018, MIT125, MIT162, SA103, NY20 and NY21 were described previously (Kasho & Katayama, 2013; Noguchi et al., 2015; Noguchi & Katayama, 2016). Strains bearing mutant oriC (NY20L and NY20ATL) were constructed using the λRed site-directed recombination system as previously described (Noguchi et al., 2015; Sakiyama et al., 2017, 2022). NY20-frt strain and oriC mutant strains (NY21-frt, NY20L-frt, and NY20ATL-frt) were constructed by eliminating kan from the NY20 strain and strains NY21, NY20L, and NY20ATL. To remove the kanamycin-resistant cassette (kan) from oriC, FLP recombinase encoded on pCP20 was used. Elimination of kan was verified by checking sensitivity to 50 μg/ml kanamycin in LB agar plates. ΔpriC::frt-kan from JW0456-KC and Δrep::frt-kan from JW5604-KC were introduced into MG1655 by P1 transduction, resulting in the construction of KRC002 and FF001, respectively. For construction of other ΔpriC::frt-kan strains (KRC004, KRC005, NY20-priC, NY21-priC, NY20L-priC, NY20ATL-priC), P1 phage lysates prepared from KRC002 were used for transduction of strains MIT125, KYA018, NY20-frt, SYM21-frt, NY20L-frt, and NY20ATL-frt. Transductants were screened on LB agar plates containing 50 µg/mL kanamycin. ΔdiaA::frt-kan or rnhA::cat derivatives (KRC006, MIT125c, and KRC004c) were constructed using a similar protocol except that P1 phage lysates from SA103 (ΔdiaA::frt-kan) or MIT162 (rnhA::cat) were used. KRC003 was generated by removal of kan from KRC002.

Buffers

Buffer P contained 60 mM HEPES–KOH (pH 7.6), 0.1 mM zinc acetate, 8 mM magnesium acetate, 30% [v/v] glycerol, and 0.32 mg/mL bovine serum albumin (BSA). Buffer N contained 50 mM HEPES–KOH (pH 7.6), 2.5 mM magnesium acetate, 0.3 mM EDTA, 7 mM DTT, 0.007% [v/v] Triton X-100, and 20% [v/v] glycerol. Form I* buffer contained 20 mM Tris-HCl (pH 7.5), 125 mM potassium glutamate, 10 mM magnesium acetate, 8 mM DTT, and 0.5 mg/mL BSA.

DUE unwinding assay

DUE unwinding assays were performed essentially as described with minor modifications (Sakiyama et al., 2022; Yoshida et al., 2023). Briefly, DnaA was incubated with ATP in buffer N on ice for 3 min to generate ATP-DnaA. The indicated amount of PriC and ATP-DnaA or its mutant derivatives was incubated for 9 min at 30℃ in 10 μL buffer P containing 5 mM ATP, 125 mM potassium glutamate, 1.6 nM pBSoriC, and 42 nM IHF, followed by further incubation with 1.5 units of P1 nuclease (Wako) for 5 min. The reaction was stopped by the addition of 1% SDS and 25 mM EDTA, and DNA was purified by phenol-chloroform extraction and ethanol precipitation. One-third of each purified DNA was digested with AlwNI (NEB), which yielded 2.6 kb and 1.0 kb fragments after DUE unwinding of pBSoriC. The resultant DNA fragments were analyzed by 1% agarose gel electrophoresis with 1xTris- acetate-EDTA buffer for 30 min at 100 V, followed by ethidium bromide staining. Gel images were taken using the GelDoc GO imaging system (Bio-Rad Laboratories, Hercules, CA), and products derived from unwound plasmids were quantified using ImageJ software.

Form I* assay

This assay was performed as previously described with minor modifications (Noguchi et al., 2015; Sakiyama et al., 2022). The indicated amount of ATP-DnaA was incubated for 15 min at 30℃ in 12.5 μL of Form I* buffer containing 3 mM ATP, 1.6 nM pBSoriC, 42 nM IHF, 400 nM His-DnaB, 400 nM DnaC, 76 nM GyrA, 100 nM His-GyrB, and 760 nM SSB. The reaction was stopped by the addition of 0.5% SDS, and DNA was purified by phenol-chloroform extraction. Samples were analyzed by 0.65% agarose gel electrophoresis with 0.5xTris-borate-EDTA buffer for 15 h at 23 V, followed by ethidium bromide staining.

Flow cytometry analysis

Flow cytometry analysis was performed essentially as described (Kasho et al., 2014). Briefly, cells were grown at 30℃ in LB medium until the absorbance of the culture (A600) reached 0.1. Portions of the cultures were diluted a thousand-fold into 5 ml LB medium and incubated at 30℃ until the absorbance of the culture (A600) reached 0.1. The remaining portions were further incubated to determine the doubling time (Td) by measuring A600 every 20 min. At A600 0.1, aliquots of the cultures were fixed in 70% ethanol to analyze cell mass using the Multisizer 3 Coulter counter (Beckman Coulter, Brea, CA). The remaining cultures were further incubated for 4 h with 0.3 mg/mL rifampicin and 0.01 mg/mL cephalexin to allow run-out replication of chromosomal DNA. The resultant cells were fixed in 70% ethanol. After DNA staining with SYTOX Green (Thermo Fisher Scientific, Waltham, MA), cellular DNA contents were analyzed on a FACS Calibur flow cytometer (BD Bioscience, Franklin Lakes, NJ). The number of the origins/cell (ori/cell) was determined from the histograms of flow cytometry analysis. The number of ori/cell was divided by the mean cell mass determined by cell mass analysis, resulting in ori/mass ratios. The numbers of cells containing non-2n copy number of oriC were divided by the number of cells containing 2n copy number of oriC to calculate the asynchrony index (A.I.). For the analysis of cells bearing plasmids, LB medium was supplemented with 100 μg/mL ampicillin.

Replication initiation frequency test using synchronized cultures

MIT125 (dnaA46, tnaA::Tn10) cells or its ΔpriC derivative KRC004 were grown at 30℃, a permissive temperature, in LB medium until the absorbance of the culture (A600) reached 0.04, followed by further incubation for 90 min at 42℃, a restrictive temperature. Aliquots of the cultures were fixed in 70% ethanol as “Synchronization” samples. The remaining cultures were further incubated for 5 min to initiate DNA replication, followed by further incubation for 4 h at 42℃ with 0.3 mg/mL rifampicin and 0.01 mg/mL cephalexin to allow run-out replication of chromosomal DNA. The resultant cells were fixed in 70% ethanol as “5 min release” samples. After DNA staining with SYTOX Green (Life Technologies), cellular DNA contents were analyzed on a FACS Calibur flow cytometer (BD Bioscience). Ratios of the origins/cell (ori/cell) were determined from the histograms of flow cytometry analysis.

For synchronization of dnaC2 mutant cells (KYA018 and KRC005), cells were grown in LB medium at 30℃ until the A600 reached 0.04, followed by further incubation for 80 min at 37℃, at the restrictive temperature. The resulting synchronized cultures were released for 5 min at 30℃ and incubated in the presence of 0.3 mg/mL rifampicin and 0.01 mg/mL cephalexin at 30℃ to allow run-out of chromosomal DNA replication.

qPCR for analysis of cSDR

Cells of MIT125 (dnaA46, tnaA::Tn10), MIT125c (dnaA46, tnaA::Tn10, rnhA::cat) and its ΔpriC derivatives KRC004 (dnaA46, tnaA::Tn10, ΔpriC::frt-kan) and KRC004c (dnaA46, tnaA::Tn10, rnhA::cat, ΔpriC::frt-kan) were grown at 30℃ in LB medium until the absorbance of the culture (A600) reached 0.04 and aliquots of the cultures were withdrawn. The remaining samples were further incubated for 90 min at 40℃ and aliquots of the cultures were withdrawn. These samples were boiled for 5 min at 95℃ and the genome DNA was extracted. The levels of oriC (84.6 min), ter (32.4 min), and yapB (50.5 min) were quantified by real-time qPCR using TB Green Premix Ex TaqII (Tli RNaseH Plus) (TaKaRa, Shiga, Japan) and the following primers: ORI_1 and KWoriCRev for oriC, qoriK fw and qoriK rev for ter, STM419 and STM420 for yapB.

eLife Assessment

This important study addresses how DNA replication restarts in Escherichia coli in the absence of a functional replication initiator protein DnaA. The authors show that helicase DnaB loading at the replication origin oriC can be executed by PriC under sub-optimal initiation conditions. While the genetic and biochemical evidence is solid, there is so far no direct evidence for PriC acting at oriC in vivo.

Reviewer #1 (Public review):

Summary:

This manuscript reports the investigation of PriC activity during DNA replication initiation in Escherichia coli. It is reported that PriC is necessary for the growth and control of DNA replication initiation under diverse conditions where helicase loading is perturbed at the chromosome origin oriC. A model is proposed where PriC loads helicase onto ssDNA at the open complex formed by DnaA at oriC. Reconstituted helicase loading assays in vitro support the model. The manuscript is well-written and has a logical narrative.

Major Questions/Comments:

An important observation here is that a ΔpriC mutant alone displays under-replication, suggesting that this helicase loading pathway is physiologically relevant. Has this PriC phenotype been reported previously? If not, would it be possible to confirm this result using an independent experimental approach (e.g. marker frequency analysis or fluorescent reporter-operator systems)?

Is PriA necessary for the observed PriC activity at oriC? Is there evidence that PriC functions independently of PriA in vivo?

Is PriC helicase loading activity in vivo at the origin direct (the genetic analysis leaves other possibilities tenable)? Could PriC enrichment at oriC be detected using chromatin immunoprecipitation?

Reviewer #2 (Public review):

This is a great paper. Yoshida et al. convincingly show that DnaA does not exclusively do loading of the replicative helicase at the E. coli oriC, but that PriC can also perform this function. Importantly, PriC seems to contribute to helicase loading even in wt cells albeit to a much lesser degree than DnaA. On the other hand, PriC takes a larger role in helicase loading during aberrant initiation, i.e. when the origin sequence is truncated or when the properties of initiation proteins are suboptimal. Here highlighted by mutations in dnaA or dnaC.

This is a major finding because it clearly demonstrates that the two roles of DnaA in the initiation process can be separated into initially forming an open complex at the DUE region by binding/nucleation onto DnaA-boxes and second by loading of the helicase. Whereas these two functions are normally assumed to be coupled, the present data clearly show that they can be separated and that PriC can perform at least part of the helicase loading provided that an area of duplex opening is formed by DnaA.

This puts into question the interpretation of a large body of previous work on mutagenesis of oriC and dnaA to find a minimal oriC/DnaA complex in many bacteria. In other words, mutants in which oriC is truncated/mutated may support the initiation of replication and cell viability only in the presence of PriC. Such mutants are capable of generating single-strand openings but may fail to load the helicase in the absence of PriC. Similarly, dnaA mutants may generate an aberrant complex on oriC that trigger strand opening but are incapable of loading DnaB unless PriC is present.

In the present work, the sequence of experiments presented is logical and the manuscript is clearly written and easy to follow. The very last part regarding PriC in cSDR replication does not add much to the story and may be omitted.

Reviewer #3 (Public review):

Summary:

At the abandoned replication fork, loading of DnaB helicase requires assistance from PriABC, repA, and other protein partners, but it does not require replication initiator protein, DnaA. In contrast, nucleotide-dependent DnaA binding at the specific functional elements is fundamental for helicase loading, leading to the DUE region's opening. However, the authors questioned in this study that in case of impeding replication at the bacterial chromosomal origins, oriC, a strategy similar to an abandoned replication fork for loading DnaB via bypassing the DnaA interaction step could be functional. The study by Yoshida et al. suggests that PriC could promote DnaB helicase loading on the chromosomal oriC ssDNA without interacting with the DnaA protein. However, the conclusions drawn from the primarily qualitative data presented in the study could be slightly overwhelming and need supportive evidence.

Strengths:

Understanding the mechanism of how DNA replication restarts via reloading the replisomes onto abandoned DNA replication forks is crucial. Notably, this knowledge becomes crucial to understanding how bacterial cells maintain DNA replication from a stalled replication fork when challenging or non-permissive conditions prevail. This critical study combines experiments to address a fundamental question of how DnaB helicase loading could occur when replication initiation impedes at the chromosomal origin, leading to replication restart.

Weaknesses:

The term colony formation used for a spotting assay could be misleading for apparent reasons. Both assess cell viability and growth; while colony formation is quantitative, spotting is qualitative. Particularly in this study, where differences appear minor but draw significant conclusions, the colony formation assays representing growth versus moderate or severe inhibition are a more precise measure of viability.

Figure 2

The reduced number of two oriC copies per cell in the dnaA46priC-deficient strain was considered moderate inhibition. When combined with the data suggested by the dnaAC2priC-deficient strain containing two origins in cells with or without PriC (indicating no inhibition)-the conclusion was drawn that PriC rescue blocked replication via assisting DnaC-dependent DnaB loading step at oriC ssDNA.

The results provided by Saifi B, Ferat JL. PLoS One. 2012;7(3):e33613 suggests the idea that in an asynchronous DnaA46 ts culture, the rate by which dividing cells start accumulating arrested replication forks might differ (indicated by the two subpopulations, one with single oriC and the other with two oriC). DnaA46 protein has significantly reduced ATP binding at 42C, and growing the strain at 42C for 40-80 minutes before releasing them at 30 C for 5 minutes has the probability that the two subpopulations may have differences in the active ATP-DnaA. The above could be why only 50% of cells contain two oriC. Releasing cells for more time before adding rifampicin and cephalexin could increase the number of cells with two oriCs. In contrast, DnaC2 cells have inactive helicase loader at 42 C but intact DnaA-ATP population (WT-DnaA at 42 or 30 C should not differ in ATP-binding). Once released at 30 C, the reduced but active DnaC population could assist in loading DnaB to DnaA, engaged in normal replication initiation, and thus should appear with two oriC in a PriC-independent manner.

Broadly, the evidence provided by the authors may support the primary hypothesis. Still, it could call for an alternative hypothesis: PriC involvement in stabilizing the DnaA-DnaB complex (this possibility could exist here). To prove that the conclusions made from the set of experiments in Figures 2 and 3, which laid the foundations for supporting the primary hypothesis, require insights using on/off rates of DnaB loading onto DnaA and the stability of the complexes in the presence or absence of PriC, I have a few other reasons to consider the latter arguments.

Figure 3

One should consider the fact that dnA46 is present in these cells. Overexpressing pdnaAFH could produce mixed multimers containing subunits of DnaA46 (reduced ATP binding) and DnaAFH (reduced DnaB binding). Both have intact DnaA-DnaA oligomerization ability. The cooperativity between the two functions by a subpopulation of two DnaA variants may compensate for the individual deficiencies, making a population of an active protein, which in the presence of PriC could lead to the promotion of the stable DnaA: DnaBC complexes, able to initiate replication. In the light of results presented in Hayashi et al. and J Biol Chem. 2020 Aug 7;295(32):11131-11143, where mutant DnaBL160A identified was shown to be impaired in DnaA binding but contained an active helicase function and still inhibited for growth; how one could explain the hypothesis presented in this manuscript. If PriC-assisted helicase loading could bypass DnaA interaction, then how growth inhibition in a strain carrying DnaBL160A should be described. However, seeing the results in light of the alternative possibility that PriC assists in stabilizing the DnaA: DnaBC complex is more compatible with the previously published data.

Figure 4

Overexpression of DiaA could contribute to removing a higher number of DnaA populations. This could be more aggravated in the absence of PriC (DiaA could titrate out more DnaA)- the complex formed between DnaA: DnaBC is not stable, therefore reduced DUE opening and replication initiation leading to growth inhibition (Fig. 4A ∆priC-pNA135). Figure 7C: Again, in the absence of PriC, the reduced stability of DnaA: DnaBC complex leaves more DnaA to titrate out by DiaA, and thus less Form I*. However, adding PriC stabilizes the DnaA: DnaBC hetero-complexes, with reduced DnaA titration by DiaA, producing additional Form I*. Adding a panel with DnaBL160A that does not interact with DnaA but contains helicase activity could be helpful. Would the inclusion of PriC increase the ability of mutant helicase to produce additional Form I*?

Figure 5

The interpretation is that colony formation of the Left-oriC ∆priC double mutant was markedly compromised at 37˚C (Figure 5B), and 256 the growth defects of the Left-oriC mutant at 25{degree sign}C and 30{degree sign}C were aggravated. However, prima facia, the relative differences in the growth of cells containing and lacking PriC are similar. Quantitative colony-forming data is required to claim these results. Otherwise, it is slightly confusing.

A minor suggestion is to include cells expressing PriC using plasmid DNA to show that adding PriC should reverse the growth defect of dnaA46 and dnaC2 strains at non-permissive temperatures. The same should be added at other appropriate places.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

This manuscript reports the investigation of PriC activity during DNA replication initiation in Escherichia coli. It is reported that PriC is necessary for the growth and control of DNA replication initiation under diverse conditions where helicase loading is perturbed at the chromosome origin oriC. A model is proposed where PriC loads helicase onto ssDNA at the open complex formed by DnaA at oriC. Reconstituted helicase loading assays in vitro support the model. The manuscript is well-written and has a logical narrative.

Thank you for understanding this study.

Major Questions/Comments:

An important observation here is that a ΔpriC mutant alone displays under-replication, suggesting that this helicase loading pathway is physiologically relevant. Has this PriC phenotype been reported previously? If not, would it be possible to confirm this result using an independent experimental approach (e.g. marker frequency analysis or fluorescent reporter-operator systems)?

We thank Reviewer 1 for this comment. This study provides the first direct evidence for PriC’s role in initiation of chromosome replication. Given the change of the oriC copy number of ∆priC cells in non-stressed conditions is only slight, resolution of the suggested methods is clearly not high enough to distinguish the differences in the oriC copy number between priC+ (WT) and ∆priC cells. Thus, to corroborate the ∆priC phenotype, we additionally analyzed using flow cytometry priC+ and ∆priC cells growing under various nutrition and thermal conditions.

As shown in Figure 2-figure supplement 1 of the revised version, the fraction of cells with non-2n oriC copies was slightly higher in ∆priC cells compared to priC+ cells. Furthermore, when grown in M9 minimal medium at 37˚C, ∆priC mutant cells exhibited slightly reduced ori/mass values. These are supportive to the idea that inhibition of replication initiation occurs at low frequency even in the WT dnaA and dnaC background, and that PriC function is necessary to ensure normal replication initiation. Related descriptions have been revised accordingly.

Is PriA necessary for the observed PriC activity at oriC? Is there evidence that PriC functions independently of PriA in vivo?

As described in Introduction of the original manuscript, PriA is a 3’-to-5’ helicase which specifically binds to the forked DNA with the 3’-end of the nascent DNA strand. Thus, structural specificity of target DNA is essentially different between PriA and PriC. Consistent with this, our in vitro data indicate that PriC alone is sufficient to rescue the abortive helicase loading at oriC (Figure 7), indicating that PriA is principally unnecessary for PriC activity at oriC. Consistently, as described in Introduction, PriC can interact with ssDNA to reload DnaB (Figure 1E). Nevertheless, a possibility that PriA might participate in the PriC-dependent DnaB loading rescue at oriC in vivo can not be completely excluded. However, elucidation of this possibility is clearly beyond the scope of the present study and should be analyzed in the future. An additional explanation has been included in Discussion of the revised version.

Is PriC helicase loading activity in vivo at the origin direct (the genetic analysis leaves other possibilities tenable)? Could PriC enrichment at oriC be detected using chromatin immunoprecipitation?

These are advanced questions about genomic dynamics of PriC. Given that PriC facilitates DnaB reloading at stalled replication forks (Figure 1E) (Heller and Marians, Mol Cell., 2005; Wessel et al., J Biol Chem, 2013; Wessel et al., J Biol Chem, 2016), PriC might interact with the whole genome and its localization might not necessarily exhibit a preference for oriC in growing cells. Analysis about these advanced questions is interesting but is beyond the scope of the present study and should be analyzed in the future study.

Reviewer #2 (Public review):

This is a great paper. Yoshida et al. convincingly show that DnaA does not exclusively do loading of the replicative helicase at the E. coli oriC, but that PriC can also perform this function. Importantly, PriC seems to contribute to helicase loading even in wt cells albeit to a much lesser degree than DnaA. On the other hand, PriC takes a larger role in helicase loading during aberrant initiation, i.e. when the origin sequence is truncated or when the properties of initiation proteins are suboptimal. Here highlighted by mutations in dnaA or dnaC.

This is a major finding because it clearly demonstrates that the two roles of DnaA in the initiation process can be separated into initially forming an open complex at the DUE region by binding/nucleation onto DnaA-boxes and second by loading of the helicase. Whereas these two functions are normally assumed to be coupled, the present data clearly show that they can be separated and that PriC can perform at least part of the helicase loading provided that an area of duplex opening is formed by DnaA. This puts into question the interpretation of a large body of previous work on mutagenesis of oriC and dnaA to find a minimal oriC/DnaA complex in many bacteria. In other words, mutants in which oriC is truncated/mutated may support the initiation of replication and cell viability only in the presence of PriC. Such mutants are capable of generating single-strand openings but may fail to load the helicase in the absence of PriC. Similarly, dnaA mutants may generate an aberrant complex on oriC that trigger strand opening but are incapable of loading DnaB unless PriC is present.

We would like to thank Revierwer#2 for the very positive comments about our work.

In the present work, the sequence of experiments presented is logical and the manuscript is clearly written and easy to follow. The very last part regarding PriC in cSDR replication does not add much to the story and may be omitted.

Given that the role PriC in stimulating cSDR was unclear, we believe that our finding that PriC has little or no role in cSDR, despite being a negative result, is valuable for the general readership of eLife. To further assess impact of PriC on cSDR and as recommended by Referee #1, we carried out the chromosome loci copy-number analysis by the whole-genome sequencing. As shown in Figure 8-supplement 1 of the revised version, the results support our conclusion from the original version.

Reviewer #3 (Public review):

Summary:

At the abandoned replication fork, loading of DnaB helicase requires assistance from PriABC, repA, and other protein partners, but it does not require replication initiator protein, DnaA. In contrast, nucleotide-dependent DnaA binding at the specific functional elements is fundamental for helicase loading, leading to the DUE region's opening. However, the authors questioned in this study that in case of impeding replication at the bacterial chromosomal origins, oriC, a strategy similar to an abandoned replication fork for loading DnaB via bypassing the DnaA interaction step could be functional. The study by Yoshida et al. suggests that PriC could promote DnaB helicase loading on the chromosomal oriC ssDNA without interacting with the DnaA protein. However, the conclusions drawn from the primarily qualitative data presented in the study could be slightly overwhelming and need supportive evidence.

Thank you for your understanding and careful comments.

Strengths:

Understanding the mechanism of how DNA replication restarts via reloading the replisomes onto abandoned DNA replication forks is crucial. Notably, this knowledge becomes crucial to understanding how bacterial cells maintain DNA replication from a stalled replication fork when challenging or non-permissive conditions prevail. This critical study combines experiments to address a fundamental question of how DnaB helicase loading could occur when replication initiation impedes at the chromosomal origin, leading to replication restart.

Thank you for your understanding.

Weaknesses:

The term colony formation used for a spotting assay could be misleading for apparent reasons. Both assess cell viability and growth; while colony formation is quantitative, spotting is qualitative. Particularly in this study, where differences appear minor but draw significant conclusions, the colony formation assays representing growth versus moderate or severe inhibition are a more precise measure of viability.

We used serial dilutions of the cell culture for the spotting assay and thus this assay should be referred as semi-quantitative rather than simply qualitative. For more quantitative assessment of viability, we analyzed the growth rates of cells and the chromosome replication activity using flow cytometry.

Figure 2

The reduced number of two oriC copies per cell in the dnaA46priC-deficient strain was considered moderate inhibition. When combined with the data suggested by the dnaAC2priC-deficient strain containing two origins in cells with or without PriC (indicating no inhibition)-the conclusion was drawn that PriC rescue blocked replication via assisting DnaC-dependent DnaB loading step at oriC ssDNA.

The results provided by Saifi B, Ferat JL. PLoS One. 2012;7(3):e33613 suggests the idea that in an asynchronous DnaA46 ts culture, the rate by which dividing cells start accumulating arrested replication forks might differ (indicated by the two subpopulations, one with single oriC and the other with two oriC). DnaA46 protein has significantly reduced ATP binding at 42C, and growing the strain at 42C for 40-80 minutes before releasing them at 30 C for 5 minutes has the probability that the two subpopulations may have differences in the active ATP-DnaA. The above could be why only 50% of cells contain two oriC. Releasing cells for more time before adding rifampicin and cephalexin could increase the number of cells with two oriCs. In contrast, DnaC2 cells have inactive helicase loader at 42 C but intact DnaA-ATP population (WT-DnaA at 42 or 30 C should not differ in ATP-binding). Once released at 30 C, the reduced but active DnaC population could assist in loading DnaB to DnaA, engaged in normal replication initiation, and thus should appear with two oriC in a PriC-independent manner.

This is a question about dnaA46 ΔpriC mutant cells. Inhibition of the replication forks causes inhibition of RIDA (the DNA-clamp complex-dependent DnaA-ATP hydrolysis) system, resulting in the increase of ATP-DnaA molecules (Kurokawa et al. (1999) EMBO J.). Thus, if ΔpriC inhibits the replication forks significantly, the ATP-DnaA level should increase and initiation should be stimulated. However, the results of Figure 2BC are opposite, indicating inhibition of initiation by ΔpriC. Thus, we infer that the inhibition of initiation in the ΔpriC cells is not related to possible changes in the ATP-DnaA level. Even if the ATP-DnaA levels are different in subpopulations in dnaA46 cells, ΔpriC mutation should not affect the ATP-DnaA levels significantly. Thus, we infer that even in dnaA46 ΔpriC mutant cells, ΔpriC mutation directly affect initiation mechanisms, rather than indirectly through the ATP-DnaA levels.

Broadly, the evidence provided by the authors may support the primary hypothesis. Still, it could call for an alternative hypothesis: PriC involvement in stabilizing the DnaA-DnaB complex (this possibility could exist here). To prove that the conclusions made from the set of experiments in Figures 2 and 3, which laid the foundations for supporting the primary hypothesis, require insights using on/off rates of DnaB loading onto DnaA and the stability of the complexes in the presence or absence of PriC, I have a few other reasons to consider the latter arguments.

This is a very careful consideration. However, we infer that stabilization of the DnaA-DnaB interaction by PriC, even if present, does not always result in stimulation of DnaB loading to oriC. Given that interactions between DnaA and DnaB during DnaB loading to oriC are highly dynamic and complicated with multiple steps, stabilization of the DnaA-DnaB interaction by PriC, even if it occurs, has a considerable risk of inhibiting the DnaB loading by constructing abortive complexes. In addition, DnaA-DiaA binding is very tight and stable (Keyamura et al., 2007, 2009). Even if WT DnaA and WT DnaB are present, PriC can rescue the initiation defects of oriC mutants. Based on these facts and the known characteristics of PriC as explained in Introduction, it is more reasonable to infer that PriC provides a bypass of DnaB loading even at oriC, as proposed for the mechanism at the stalled replication fork. However, we cannot completely rule out the indicated possibility and these explanations are included in the revised version.

Figure 3

One should consider the fact that dnA46 is present in these cells. Overexpressing pdnaAFH could produce mixed multimers containing subunits of DnaA46 (reduced ATP binding) and DnaAFH (reduced DnaB binding). Both have intact DnaA-DnaA oligomerization ability. The cooperativity between the two functions by a subpopulation of two DnaA variants may compensate for the individual deficiencies, making a population of an active protein, which in the presence of PriC could lead to the promotion of the stable DnaA: DnaBC complexes, able to initiate replication. In the light of results presented in Hayashi et al. and J Biol Chem. 2020 Aug 7;295(32):11131-11143, where mutant DnaBL160A identified was shown to be impaired in DnaA binding but contained an active helicase function and still inhibited for growth; how one could explain the hypothesis presented in this manuscript. If PriC-assisted helicase loading could bypass DnaA interaction, then how growth inhibition in a strain carrying DnaBL160A should be described. However, seeing the results in light of the alternative possibility that PriC assists in stabilizing the DnaA: DnaBC complex is more compatible with the previously published data.

Unfortunately, in this comment, there is a crucial misunderstanding in the growth of cells bearing DnaA L160A. Hayashi et al. reported that the dnaB(Ts) cells bearing the dnaB L160A allele grew slowly and formed colonies even at 42°C. This feature is similar to the growth of dnaA46 cells bearing dnaA F46A H136A allele (Figure 2). Thus, the results of dnaB L160A cells are consistent with our model and support the idea that PriC partially rescues the growth inhibition of cells bearing the DnaB L160A allele by bypassing the strict requirement for the DnaA-DnaB interaction. Nevertheless, we have to be careful about a possibility that DnaB L160A could affect interaction with PriC, which we are going to investigate for a future paper.

As suggested, if mixed complexes of DnaA46 and DnaA F46A H136A proteins are formed, those might retain partial activities in oriC unwinding and DnaB interaction although those cells are inviable at 42°C without PriC. It is noteworthy that in the specific oriC mutants which are impaired in DnaB loading (e.g., Left-oriC), PriC effectively rescues the initiation and cell growth. In these cells, both DnaA and DnaB are intact. Thus, the idea that only mutant DnaA (or DnaB) protein is simulated specifically via PriC interaction is invalid. Even in cells bearing wild-type oriC, DnaA and DnaB, contribution of PriC for initiation is detected.

In addition, as described in the above response, given that interactions between DnaA and DnaB during DnaB loading to oriC are very dynamic and complicated with multiple steps, stabilization of the DnaA-DnaB interaction by PriC, even if present, would not simply result in stimulation of DnaB loading to oriC; rather we think a probability that it would inhibit the DnaB loading by constructing abortive complexes. Based on the known characteristics of PriC as explained in Introduction, it is more reasonable to infer that PriC provides a bypass of DnaB loading even at oriC, as proposed for the mechanism at the stalled replication fork.

However, we cannot completely rule out the indicated possibility and this explanation has been described in the revised version as noted in response to the above question.

Figure 4

Overexpression of DiaA could contribute to removing a higher number of DnaA populations. This could be more aggravated in the absence of PriC (DiaA could titrate out more DnaA)-the complex formed between DnaA: DnaBC is not stable, therefore reduced DUE opening and replication initiation leading to growth inhibition (Fig. 4A ∆priC-pNA135). Figure 7C: Again, in the absence of PriC, the reduced stability of DnaA: DnaBC complex leaves more DnaA to titrate out by DiaA, and thus less Form I*. However, adding PriC stabilizes the DnaA: DnaBC hetero-complexes, with reduced DnaA titration by DiaA, producing additional Form I*. Adding a panel with DnaBL160A that does not interact with DnaA but contains helicase activity could be helpful. Would the inclusion of PriC increase the ability of mutant helicase to produce additional Form I*?

Unfortunately, the proposed idea is biased disregarding the fact that DiaA effectively stimulates assembling processes of DnaA molecules at oriC. As oriC contains multiple DnaA boxes and multiple DnaA molecules are recruited there, DiaA will efficiently facilitate assembling of DnaA molecules on oriC. Even DnaA molecules of DnaA-DiaA complexes can efficiently bind to oriC. This is consistent with in vitro experiments showing that higher levels of DiaA stimulate assembly of DnaA molecules and oriC unwinding (i.e., DUE opening) but even excessive levels of DiaA do not inhibit those reactions (Keyamura et al., J. Biol. Chem. (2009) 284, 25038-25050). However, as shown in Figure 9, DiaA tightly binds to the specific site of DnaA which is the same as the DnaB L160-binding site, which causes inhibition of DnaA-DnaB binding (ibid). These are consistent with in vivo experiments, and concordantly consistent with the idea that the excessive DiaA level inhibits interaction and loading of DnaB by the DnaA-oriC complexes, but not oriC unwinding (i.e., DUE opening) in vivo. Also, as mentioned above, we do not consider that stabilization of DnaA-DnaBC complex simply results in stimulation of DnaB loading to oriC. Based on the known characteristics of PriC, it is more reasonable to infer that PriC provides a bypass of DnaB loading even at oriC, as proposed for the mechanism at the stalled replication fork (Figure 1E), as described in the above response.

As for DnaB L160A, as mentioned above, we are currently investigating interaction modes between DnaB and PriC. While investigating DnaB L160A could further support our model, we believe its contribution to the present manuscript would be incremental. In addition, there is a possibility that DnaA L160A could affect interaction with PriC. Thus, analysis of DnaB mutants in this PriC rescue mechanisms should be addressed in future study.

Figure 5

The interpretation is that colony formation of the Left-oriC ∆priC double mutant was markedly compromised at 37˚C (Figure 5B), and 256 the growth defects of the Left-oriC mutant at 25{degree sign}C and 30{degree sign}C were aggravated. However, prima facia, the relative differences in the growth of cells containing and lacking PriC are similar. Quantitative colony-forming data is required to claim these results. Otherwise, it is slightly confusing.

The indicated concern was raised due to our typing error lacking ∆priC. In the revised manuscript, we have amended as follows: the cell growth of the Left-oriC ∆priC double mutant was markedly compromised at 37˚C and moderately reduced at 25°C and 30°C (Figure 5B).

A minor suggestion is to include cells expressing PriC using plasmid DNA to show that adding PriC should reverse the growth defect of dnaA46 and dnaC2 strains at non-permissive temperatures. The same should be added at other appropriate places.

Even in the presence of PriC, unwinding of oriC and DnaB helicase loading to the wound oriC require DnaA and DnaC activities as indicated by previous studies (see for a review, Windgassen et al., (2018) Nucleic Acids Res. 46, 504-519). Thus, dnaA46 cells and dnaC2 cells bearing pBR322-priC can not grow at 42°C and 37°C (as follows). These are reasonable results. However, at semi-permissive temperatures (37°C for dnaA46 and 35°C for dnaC2), slight stimulation of the cell growth by pBR322-priC might be barely observed (Figure 2-supplement 1 of the revised version). These suggest that the intrinsic level of PriC is functionally nearly sufficient. This explanation has been included in the revised version.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Line 38. "in assembly of the replisome".

Corrected.

Line 137. "specifically" rather than specificity.

Corrected.

Line 139. "at" rather than by.

Corrected.

The DnaA46 protein variant contains two amino acid substitutions (A184V and H252Y) within the AAA+ motif. H136 appears to reside adjacent to A184 in structure. Is A184V mutation causative?

The DnaA H136A and A184V alleles are responsible for different defects. Indeed, the DnaA A184V variant is thermolabile and defective in ATP binding whereas the H136A variant retains ATP binding but impairs DnaB loading (Carr and Kaguni, Mol. Microbiol., 1996; Sakiyama et al., Front. Microbiol., 2018). These observations strongly support the view that the phenotype of the DnaA H136A allele is independent of that of the DnaA A184V allele.

Figure 2A. Regarding the dnaA46 allele grown at 37°C.

Individual colonies cannot be resolved. Is an image from a later time-point available?

We have replaced the original image with one from another replicate that provides better resolution. Please see Figure 2A in the revised version.

Figure 2C. Quantification of the number of cells with more than one chromosome equivalent in the dnaC2 ΔpriC strain. The plot from flow cytometry appears to show >20% of cells with only 1 genome. Are these numbers correct?

Thank you for this careful comment. We quantified the peaks more strictly, but the percentages were noy largely changed. To improve resolution of the DNA profiles, we have changed the range of the x-axis in panels B and C of Figure 2 in the revised version.

Figure 3. Are both F46A and H136A mutations in the plasmid-encoded dnaA necessary?

Yes. The related explanation is included in the Discussion section (the third paragraph) of the original manuscript. As described there, dnaA46 cells expressing the DnaA H136A single mutant exhibited severe defects in cell growth even in the presence of PriC (Sakiyama et al., 2018). The His136 residue is located within the weak, secondary DnaB interaction region in DnaA, and is crucial for DnaB loading onto oriC ssDNA. Given domain I in DnaA H136A can stably tether DnaB-DnaC complexes to DnaA complexes on oriC (Sakiyama et al., 2018), we infer that oriC-DnaA complexes including DnaA H136A stably bind DnaB via DnaA domain I as an abortive complex, which inhibits functional interaction between PriC and DnaB as well as DnaB loading to oriC DNA.

As for DnaA F46A mutant, our previous studies show that DnaA F46A has a limited residual activity in vivo (unlike in vitro), and allows slow growth of cells. As the stable DnaA-DnaB binding is partially impaired in vivo in DnaA F46A, this feature is consistent with the above ideas. Thus, both F46A and H136A mutations are required for severer inhibition of DnaB loading. This is additionally described in the revised Discussion.

Figure 3. Is the DnaA variant carrying F46A and H136A substitutions stably expressed in vivo?

We have performed western blotting, demonstrating that the DnaA variant carrying F46A and H136A substitutions is stable in vivo. In the revised version, we have added new data to Figure 3-figure supplement 1 and relevant description to the main text as follows:

Western blotting demonstrated that the expression levels were comparable between WT DnaA and DnaA F46A H136A double mutant (Figure 3-figure supplement 1).

Figure 5A. Should the dashed line extending down from I2 reach the R4Tma construct?

We have amended the indicated line appropriately.

Figure 6C. It was surprising that the strain combining the subATL mutant with ΔpriC displayed a pronounced under-initiation profile by flow cytometry, and yet there was no growth defect observed (see Figure 6B). This seems to contrast with results using the R4Tma origin, where the ΔpriC mutant produced a relatively modest change to the flow cytometry profile, and yet growth was perturbed (Figure 5C-D). How might these observations be interpreted? Is the absolute frequency of DNA replication initiation critical?

Please note that, in E. coli, initiation activity corelates closely with the numbers of oriC copies per cell mass (ori/mass), rather than the apparent DNA profiles measured by flow cytometer. When cells were grown in LB at 30˚C, the mean ori/mass values were as follows: 0.34 for R4Tma priC, 0.51 for R4Tma, 0.82 for DATL priC, 0.99 for DATL (Figures 5 & 6 in the original manuscript). These values closely correspond to the cell growth ability shown in Figure 5C in the original manuscript.

In the revised manuscript, we have cited appropriate references for introduction of the ori/mass values as follows.

To estimate the number of oriC copies per unit cell mass (ori/mass) as a proxy for initiation activity (Sakiyama et al., 2017, 2022),

Line 295. Reference for Form I* assay should cite the original publication.

Done. The following paper is additionally cited.

Baker, T. A., Sekimizu, K., Funnell, B. E., and Kornberg, A. (1986). Extensive unwinding of the plasmid template during staged enzymatic initiation of DNA replication from the origin of the Escherichia coli chromosome. Cell 45, 53–64.doi: 10.1016/0092-8674(86)90537-4

Reviewer #2 (Recommendations for the authors):

The partial complementation of the dnaC2 strain by PriC seems quite straightforward since this particular mutation leads to initiation arrest at the open complex stage and this sets the stage for PriC to load the helicase. The situation is somewhat different for dnaA46. Why is this mutation partly complemented by PriC at 37C? DnaA46 binds neither ATP nor ADP, yet it functions in initiation at permissive temperature. At nonpermissive temperature, it binds oriC as well but does not lead to initiation. Does the present data imply that the true initiation defect of DnaA46 lies in helicase loading? The authors need to comment on this in the text.

Given the thermolabile propensity of the DnaA46 protein, it is presumable that DnaA46 protein becomes partially denatured at the sub-permissive temperature of 37˚C. This partial denaturation should impair both origin unwinding and helicase loading, though not to the extent that cell viability is lost. The priC deletion should further exacerbate helicase loading defects by inhibiting the bypass mechanism, resulting in the lethality of dnaA46 cells at this temperature. This explanation is included in the revised Discussion section.

Relating to the above. In Figure 3 it is shown that the pFH plasmid partly complements dnaA46 in a PriC-dependent manner. Again, it would be nice to know the nature of the DnaA46 protein defect. It would be interesting to see how a pING1-dnaA46 plasmid performs in the experiment presented in Figure 3.

A previous paper showed that multicopy supply of DnaA46 can suppress temperature sensitivity of the dnaA46 cells (Rao and Kuzminov, G3, 2022). This is reasonable in that DnaA46 has a rapid degradation rate unlike wild-type DnaA. As DnaA46 preserves the intact sequences in DnaB binding sites such as G21, F46 and H136, the suppression would not depend on PriC but would be due to the dosage effect.

Figure 8 B: The authors should either remove the data or show a genome coverage: it is not clear that yapB is a good reference. A genome coverage would be nice, and show whether initiation can occur at oriC even if it is not the major place of initiation in a rnhA mutant.

As suggested, we carried out the chromosome loci copy-number analysis by whole-genome sequencing to assess impact of PriC on cSDR. The new data are shown in Figure 8-supplement 1 with relevant descriptions of the main text of the revised version as shown below. Briefly, results of the chromosome loci copy-number analysis are consistent with those of real-time qPCR (Figure 8B). Given that the role PriC in stimulating cSDR was unclear, we believe that our finding that PriC has little or no role in cSDR, despite being a negative result, is valuable for the general readership of eLife.

Line 38-39: .....resulting in replisome assembly.

Corrected.

Line 48: Something is wrong with the Michel reference. Also in the reference list.

Corrected

Line 156: replace retarded with reduced.

Corrected.

Line 171 and elsewhere: WT priC cells is somewhat misleading. Isn't this simply PriC+ cells?

Yes. We have revised the wording to “priC+” for clarity.

Line 349-350: "the oriC copy number ratio of the dnaA46 DpriC double mutant was lower than that of the dnaA46 single mutant....". This is only provided growth rate of the strains is the same.

These strains exhibited similar growth rates. This is included in the Result section of the revised manuscript as follows: At the permissive temperature, despite having similar growth rates, the oriC copy number ratio of the dnaA46 ∆priC double mutant strain was lower than that of the dnaA46 single mutant.

Reviewer #3 (Recommendations for the authors):

I would suggest improved or additional experiments, data, or analyses.

The revised version includes improved or additional experiments, data, or analyses.

Version 2

Abstract

Abstract

In Escherichia coli, replisome and replication fork assembly is initiated by DnaB helicase loading at the chromosomal origin oriC via its interactions with the DnaA initiator and the DnaC helicase loader. Upon replication fork arrest, the replisome including DnaB dissociates from the stalled fork. Replication fork progression is rescued by primosomal protein PriA- or PriC-dependent pathway in which PriA and PriC promote reloading of DnaB in different mechanisms. However, the mechanism responsible for rescue of blocked replication initiation at oriC remains unclear. Here, we found that PriC rescued blocked replication initiation in cells expressing an initiation-specific DnaC mutant, in mutant cells defective in DnaA-DnaB interactions, and in cells containing truncated oriC sequence variants. PriC rescued DnaB loading at oriC even in the absence of Rep helicase, a stimulator of the PriC-dependent replication fork restart pathway. These and results of in vitro reconstituted assays concordantly suggest that this initiation-specific rescue mechanism provides a bypass of the DnaA-DnaB interaction for DnaB loading by PriC-promoted loading of DnaB to the unwound oriC region. These findings expand understanding of mechanisms sustaining the robustness of replication initiation and specific roles for PriC in the genome maintenance.

Introduction

Chromosome replication is strictly regulated to ensure successful duplication of the genetic material (Costa & Diffley, 2022; Grimwade & Leonard, 2021; Kasho et al., 2023). In Escherichia coli, the ATP-bound form of the replication initiator protein DnaA (ATP-DnaA) forms a higher-order complex with DNA-bending protein IHF at the unique chromosomal origin oriC. This complex promotes local unwinding of oriC and recruits a pair of DnaB helicases, which are successively loaded to the single-stranded (ss) DNA regions in a bidirectional manner with the aid of DnaC helicase loader. ssDNA-loaded DnaB interacts with DnaG primase and DNA polymerase III holoenzyme, resulting in assembly of the replisome (Arias-Palomo et al., 2019; Chodavarapu & Kaguni, 2016; Hayashi et al., 2020; O’Donnell et al., 2013; Sakiyama et al., 2017, 2018, 2022; Wegrzyn & Konieczny, 2023).

Replisomes disassemble when the replication fork stalls in front of an obstacle, such as protein-DNA roadblock or single strand breaks. PriA recognizes the abandoned fork structure, which triggers reloading of DnaB with its partner proteins, such as PriB, PriC, DnaT, and DnaC (Heller & Marians, 2005a; Masai et al., 1994; Michel & Sandler, 2017; Windgassen et al., 2018). In addition, independently of PriA, PriC binds to the abandoned fork structure, triggering reloading of DnaB with the aid of DnaC, which is supported by Rep helicase depending on the fork structure (Heller & Marians, 2005b; Sandler et al., 1999). Also, homologous recombination mechanisms can rescue the abandoned replication fork during SOS responses (Asai & Kogomal, 1994; Kogoma, 1997; Né Dicte Michel et al., 2001). In contrast to these well characterized mechanisms at abandoned forks, little is known about what happens when replication initiation is impeded at oriC.

E. coli oriC comprises a Duplex-Unwinding Element (DUE) and a DnaA-Oligomerization Region (DOR) (Figure 1A) (Grimwade & Leonard, 2021; Kasho et al., 2023; Wegrzyn & Konieczny, 2023). The DUE contains 13-mer AT-rich sequence repeats known as L-, M-, and R-DUEs. The M- and R-DUEs are essential for stable DUE unwinding with the specific sequences (TT[A/G]T(T)) used for DnaA-ssDNA binding (Ozaki & Katayama, 2012; Sakiyama et al., 2017). The L-DUE containing the TTATT sequence promotes efficient DnaB loading by expanding the unwound oriC region (Sakiyama et al., 2022). The DOR is divided into three subregions: the Left-, Middle-, and Right-DORs, which contain asymmetric 9-mer DnaA binding sequences (DnaA box) with the consensus sequence TTA[T/A]NCACA (Figure 1A) (Noguchi et al., 2015; Ozaki & Katayama, 2012; Rozgaja et al., 2011; Sakiyama et al., 2017; Shimizu et al., 2016). The Left-DOR contains a cluster of unidirectionally aligned DnaA boxes including high-affinity DnaA box R1, and low-affinity boxes R5M, τ1−2, and I1-2, with an IHF-binding region between R1 and R5M boxes. The Right-DOR contains an oppositely oriented DnaA box cluster including high-affinity DnaA box R4 and low-affinity boxes I3 and C1-3. These DnaA box clusters form frameworks for the Left- and Right-DnaA subcomplexes, respectively. DnaA bound to the R2 box, which solely resides in the Middle-DOR, stabilizes these DnaA subcomplexes (Miller et al., 2009; Rozgaja et al., 2011; Shimizu et al., 2016). In addition, the AT-cluster (TATTAAAAAGAA) region, which connects to the L-DUE, stimulates DnaB loading in the absence of the Right-DnaA subcomplex (Sakiyama et al., 2022).

Schematic structures of oriC and DnaA and the replication initiation mechanism. (A) Overall structure of oriC with the AT-rich cluster. oriC (245 bp) includes the Duplex-Unwinding Element (DUE, purple bar) and the DnaA-Oligomerization Region (DOR, gray bar). DUE is composed of three AT-rich 13-mer repeats termed L, M, and R. Dark purple boxes indicate the specific sequences involved in DnaA-ssDNA binding (TT[A/G]T(T)). The AT-cluster (purple dots) flanking the DUE outside oriC is a supplementary unwinding region. The DOR contains 12 DnaA boxes (filled and open arrowheads representing sites with the full consensus TTA[A/T]NCACA sequence and sites with mismatches, respectively) and an IHF-binding region (IBR, green box). The DOR is subdivided into Left, Middle, and Right subregions. (B) Domains of DnaA. Domains I-IV are shown schematically, with amino acid residue numbers shown in each bracket. H/B motifs (V211 and R245, squares) and Arginine finger (R285, triangle) are indicated. The major functions of each domain are described on the right side of the structure. (C) DUE unwinding by the ssDUE recruitment mechanism. DnaA and IHF are indicated by red and green diagrams, respectively. When ATP-DnaA oligomerizes, DUE is unwound unstably by thermal motion and torsional stress. The M/R region of the DUE-upper strand (purple line) binds to R1-DnaA and R5M-DnaA via IHF-induced DNA bending, resulting in stable DUE unwinding. In addition to M/R-DUE, L-DUE is moderately unwound and interacts with the Right-DnaA subcomplex. (D) Mechanism for DnaB loading. Each Left-DnaA and Right-DnaA subcomplex binds to a DnaB hexamer (blue ovals) complexed with DnaC (yellow circles). (E) PriC-dependent fork rescue pathways. PriC (vivid red circle) promotes DnaB reloading to the abandoned fork, which exposes ssDNA regions. PriC binds to ssDNA at the replication fork, which remodels the SSB-ssDNA complex (for simplicity, SSB is omitted) for DnaC-dependent reloading of DnaB (upper panel). If the ssDNA region is accompanied with a nascent lagging strand, Rep helicase (green shape) is recruited by PriC and unwinds the lagging strand to expose the ssDNA region for DnaB loading (the lower case).

DnaA comprises four functional domains (Figure 1B). Domain I binds to multiple proteins such as DnaB helicases and the DnaA-assembly stimulator DiaA (Abe et al., 2007; Hayashi et al., 2020; Keyamura et al., 2007, 2009). Domain II is a flexible linker (Abe et al., 2007; Nozaki & Ogawa, 2008). Domain III contains AAA+ (ATPases Associated with various cellular Activities) motifs, which are involved in tight ATP/ADP binding, ATP hydrolysis, and domain III-III interactions (Duderstadt et al., 2011; Erzberger et al., 2006; Felczak & Kaguni, 2004; Kawakami et al., 2005; Ozaki et al., 2012). The arginine finger motif (Arg285) in domain III interacts with ATP bound to domain III of adjacent DnaA protomers, stimulating DnaA complex formation in a cooperative manner (Kawakami et al., 2005; Noguchi et al., 2015). In addition, H/B motifs (Val 211 and Arg245) in this domain bind to ssDNA in a sequence-specific manner (Ozaki et al., 2008; Sakiyama et al., 2017). Domain IV, which comprises the helix-turn-helix motif, recognizes the DnaA box (Fujikawa et al., 2003).

For replication initiation, ATP-DnaA molecules cooperatively oligomerize at oriC to form the Left- and Right-DnaA subcomplexes (Figure 1C). Formation of these complexes is stimulated by homotetrameric DiaA protein, which binds up to four DnaA molecules via domain I, including Phe46 (Keyamura et al., 2007, 2009). In concert with DNA superhelicity and thermal energy, the DUE undergoes initial unstable unwinding, which is stabilized through specific interaction between H/B motifs of domain III in the Left-DnaA subcomplex and the upper strand of M- and R-DUEs. This interaction is facilitated by sharp DNA bending by IHF, resulting in stable DUE unwinding (Figure 1C) (Ozaki et al., 2008; Sakiyama et al., 2017). When the single-stranded region is expanded to the L-DUE, the resultant single-stranded L-DUE binds the H/B motifs within the Right-DnaA subcomplex, maximizing the efficiency of the DnaB loading process (Sakiyama et al., 2022).

Subsequently, two DnaB helicases are loaded onto each ssDUE strand through interactions with DnaC and DnaA (Figure 1D). DnaC binding to DnaB changes the closed ring structure of the DnaB hexamer to an open spiral form to allow it to encircle ssDNA (Arias-Palomo et al., 2019; Nagata et al., 2020). Each Left- and Right-DnaA subcomplex binds a DnaB-DnaC complex via high-affinity interaction between DnaA Phe46 and DnaB Leu160 and guides the DnaB-DnaC complexes to unwound ssDUE strands via a low-affinity interaction between DnaA domain III His136 and DnaB (Figure 1B and 1D). Upon interaction with ssDNA, DnaC dissociates from DnaB, which enables migration of DnaB hexamer with helicase action. The pair of DnaB helicases loaded onto the ssDNA strands progress in opposite directions, forming replisomes with DnaG and Pol III holoenzyme (Chodavarapu & Kaguni, 2016; O’Donnell et al., 2013).

Replication restart pathways ensure fork progression of the entire chromosome under conditions that trigger replisome disassembly (Heller & Marians, 2005a; Lopper et al., 2007; Michel & Sandler, 2017). PriA helicase-dependent pathways predominantly facilitate DnaB reloading onto abandoned forks in vivo (Flores et al., 2002). PriA is a 3’-5’ DNA helicase and has specific affinity for forked DNA structures comprising one parental dsDNA and two newly synthesized sister dsDNA strands. This permits PriA recognition of the abandoned replication fork and the subsequent unwinding of the nascent lagging strand (Duckworth et al., 2023; Windgassen et al., 2018; Windgassen & Keck, 2016). The resultant unwound ssDNA associates with PriB and DnaT to promote reloading of DnaB helicase with the aid of DnaC (Duckworth et al., 2023; Heller & Marians, 2005a; Lopper et al., 2007). In the absence of PriB, PriC participates in a PriA-dependent pathway (Sandler et al., 1999). In addition, PriA is required for oriC/DnaA-independent chromosomal replication called stable DNA replication (SDR) (Masai et al., 1994), which is promoted by UV irradiation (inducible iSDR) and loss of rnhA or recG (constitutive cSDR) (Kogoma, 1997). In both mechanisms, replication is initiated at triple-stranded structures comprising a dsDNA-ssDNA hybrid (D-loop) or a dsDNA-ssRNA hybrid (R-loop) generated at specific chromosomal loci. These structures mimic the abandoned fork structure, thereby allowing helicase loading to occur similarly to that during PriA-dependent fork restart (Masai et al., 1994).

Independently of PriA, PriC can restart replication from abandoned forks through its interaction with the ssDNA region and SSB (Single-Stranded DNA Binding protein) (Figure 1E) (Heller & Marians, 2005a; Wessel et al., 2013, 2016). PriC consists of N-terminal and C-terminal domains, which are composed of α helices and are connected by a short linker. The PriC C-terminal domain remodels the SSB-ssDNA complex, which is a prerequisite for the recruitment of DnaB helicase (Heller & Marians, 2005a; Wessel et al., 2013, 2016). PriC also interacts with Rep helicase. When the length of the ssDNA gap within the abandoned fork is short, Rep helps expand the ssDNA gap, promoting PriC-mediated remodeling of the SSB-ssDNA complex (Heller & Marians, 2005a, 2005b; Nguyen et al., 2021). Notably, in addition to these functions at the abandoned forks, PriC is suggested to play a role in DnaA-oriC-dependent replication initiation under challenging conditions, as evidenced by the synthetic lethality between priC303::kan and a subgroup of temperature-sensitive dnaA mutant alleles (dnaA46 and dnaA508): i.e. priC303::kan mutants bearing dnaA46 or dnaA508, but not wild-type (WT) dnaA, can not grow even at 30°C (Hinds & Sandler, 2004). However, the mechanisms via which PriC rescues blocked replication initiation remain elusive.

In this study, we analyzed the PriC-dependent promotion of replication initiation from oriC in the context of various replication initiation stresses. We found that PriC stimulated replication initiation in dnaC2 cells, in which DnaB helicase loading at oriC is defective, indicating that it has a role in the rescue of blocked replication initiation. Moreover, PriC stimulated the growth of cells defective in DnaA-DnaB interactions in addition to the growth of cells with oriC sequence deletions that inhibit DnaB loading. Consistent with these results, in an in vitro reconstituted system, PriC stimulated DnaB loading when the DnaA-DnaB interaction was inhibited at oriC. Furthermore, we found that PriC did not stimulate initiation of cSDR, demonstrating that PriC functions specifically in the DnaA-oriC system. Taken together, we suggest that PriC rescues blocks in replication initiation by bypassing DnaB loading at oriC-DnaA complexes.

Results

PriC stimulates DNA replication initiation in dnaA46 and dnaC2 cells

The temperature-sensitive dnaA46 and dnaA508 mutants exhibit synthetic lethality with priC303::kan even at 30°C (Hinds & Sandler, 2004), suggesting that PriC can rescue blocks in DnaA-dependent replication initiation. However, the mechanism underlying PriC-promoted rescue of blocks in replication initiation remains a mystery. Based on previous reports demonstrating PriC binding to short oligo-ssDNA, SSB, and DnaB and its role in DnaB loading at stalled replication forks (Wessel et al., 2013, 2016), we considered the possibility that PriC permits bypass of DnaA-dependent stable DUE unwinding and/or DnaB loading at the oriC ssDUE. To genetically test these ideas, we analyzed the impact of PriC on the growth of a temperature-sensitive dnaC2 mutant, which is competent for ongoing replisome progression at 37–42°C, but defective in DnaB loading specifically at the oriC ssDUE (Withers & Bernander, 1998).

In spotting assays performed using serial dilutions of the cell cultures and LB agar plates, the cell growth of dnaA46 cells was slightly inhibited at 37℃ (Figure 2A). Consistent with the previous report, when crossed with the ΔpriC allele, the cell growth of dnaA46 cells but not WT dnaA cells was severely inhibited at 37℃ (Figure 2A). Unlike the dnaA46 priC303::kan cells, the dnaA46 ΔpriC cells grew at 30°C without prominent inhibition (Figure 2A). This difference may stem from the different strain backgrounds. Notably, the cell growth of dnaC2 mutant at 35℃ was moderately inhibited, whereas it was severely inhibited in dnaC2 ΔpriC double mutant (Figure 2A). Also, the cell growth of dnaC2 cells and dnaC2 ΔpriC cells was similar at 30°C. The observed requirement of PriC for cell growth of dnaC2 cells at 35℃ supports the idea that PriC assists in the DnaC-dependent DnaB loading step at oriC.

Requirement of priC for DNA replication initiation in dnaA46 and dnaC2 strains. (A) Cell growth abilities. MG1655 and its derivatives MIT125 (MG1655 dnaA46 tnaA::Tn10), KYA018 (MG1655 dnaC2 zjj18::cat), KRC002 (MG1655 ΔpriC::frt-kan), KRC004 (MG1655 dnaA46 tnaA::Tn10 ΔpriC::frt-kan), and KRC005 (MG1655 dnaC2 zjj18::cat ΔpriC::frt-kan) were grown overnight. 10-fold-serial dilutions of the overnight cultures (∼109 cells/mL) were incubated for 16 h at 30℃, or for 14 h at 35/37/42℃ on LB agar medium. Three independent experiments were performed. +, WT priC; -, ΔpriC::frt-kan. The images of representative plates are shown. (B) DNA replication initiation in synchronized dnaA46 cells with or without priC. Exponentially growing MIT125 (MG1655 dnaA46 tnaA::Tn10) (+) and KRC004 (MG1655 dnaA46 tnaA::Tn10 ΔpriC::frt-kan) (-) cells were synchronized to the pre-initiation stage by incubation for 90 min at 42℃ in LB medium, followed by incubation for 5 min at 30℃ to initiate DNA replication, which was further followed by incubation with rifampicin and cephalexin for 4 h at 42℃ to allow run-out of ongoing DNA replication. Samples were withdrawn after synchronization and run-out replication. The DNA contents of these samples were measured by flow cytometry. Representative histograms of three independent experiments are shown. For each sample, the mean number of cells with more than one chromosome equivalent after a 5 min release was quantified and is shown as a percentage with standard deviations. (C) DNA replication initiation in synchronized dnaC2 cells with or without priC. Exponentially growing KYA018 (MG1655 dnaC2 zjj18::cat) (+) and KRC005 (MG1655 dnaC2 zjj18::ca ΔpriC::frt-kan) (-) cells were synchronized to the pre-initiation stage by incubation for 80 min at 37℃ in LB medium, followed by incubation for 5 min at 30℃ and further incubation with rifampicin and cephalexin as described above. Samples were withdrawn and analyzed as described above.

Introduction of a priC-encoded plasmid (pBRpriC), but not an empty vector (pBR322), restored the cell growth of dnaA46 ΔpriC double mutant at 37℃ (Figure 2-figure supplement 1AB) and that of dnaC2 ΔpriC double mutant at 35℃. Also, introduction of pBRpriC to dnaA46 and dnaC2 mutants with intact priC (priC+) might only slightly enhance the cell growth at these temperatures (Figure 2-figure supplement 1AB), suggesting that the intrinsic level of PriC is functionally nearly sufficient.

To investigate replication initiation from oriC, we used flow cytometry of synchronized cells bearing dnaA46 or dnaC2 with or without ΔpriC. In these experiments, cells were incubated at 42°C or 37°C for 80–90 min to synchronize the replication cycle at the pre-initiation step, after which incubation was continued at 30°C for 5 min to allow initiation and then 4 h in the presence of rifampicin and cephalexin to inhibit both replication initiation and cell division and allow replication run-out of the whole chromosome. The resultant number of chromosomes per cell, measured by flow cytometry, is known to correspond to the number of oriC copies per cell at the time of drug addition, an indicator of initiation activity.

A majority of the synchronized dnaA46 and dnaC2 cells had only a single oriC copy per cell after incubation at 42°C or 37°C regardless of the presence of ΔpriC (Figure 2B and 2C). In cells with intact priC (priC+), after release of initiation at 30°C for 5 min, most cells had two oriC copies.

However, in dnaA46 ΔpriC cells, only about half of the cells had two oriC copies after the release (Figure 2B), indicating moderate inhibition of initiation. In dnaC2 ΔpriC cells, the number of two-oriC cells was slightly lower than that of dnaC2 cells (Figure 2C). These observations are consistent with the cell growth ability (Figure 2A) and with our hypothesis that PriC contributes to the rescue of blocked replication initiation by assisting DnaB loading onto the oriC ssDNA.

We further performed flow cytometry analysis to assess the role for PriC in replication initiation of the WT dnaA/dnaC cells. To estimate the number of oriC copies per unit cell mass (ori/mass) as a proxy for initiation activity (Sakiyama et al., 2017, 2022), exponentially growing cells were subjected to run-out replication in the presence of rifampicin and cefalexin (Figure2-figure supplement 2). When cells were grown in LB medium at 25°C, 30°C or 42°C, the ori/mass values for ΔpriC mutant were comparable to those of priC+ cells. Similar results were observed in cells grown in M9 minimal medium at 30°C. However, when grown in M9 minimal medium at 37°C, ΔpriC mutant cells exhibited slightly reduced ori/mass values. Also, it is noteworthy that in all conditions we tested, the fraction of cells with non-2n oriC copies was slightly higher in ΔpriC cells compared to priC+cells. The asynchronous initiations would be caused by partial inhibition of initiation of multiple origins in ΔpriC cells, as representatively seen for the case of ΔpriC cells growing in M9 minimal medium at 37°C where four-oriC cells reduced and three-oriC cells increased, compared to priC+ cells. A similar case might be shown for ΔpriC cells growing in LB medium at 25°C in that eight-oriC cells reduced and asynchronous initiations (A.I. %) as well as four-oriC cells increased. As such, inhibition of replication initiation would occur intrinsically at low frequency in the WT dnaA/dnaC background. PriC function would be effective to suppress the inhibition, thereby ensuring replication initiation of multiple origins.

PriC rescues the cell growth of DnaA mutants defective in DnaB interactions

DnaA contains two regions for interactions with DnaB during DnaB loading at oriC ssDNA: one is a high-affinity region in domain I containing Phe46, and the other is a low-affinity region in domain III containing His136. To analyze the mechanism of PriC-promoted DnaB loading at oriC, we introduced a series of pING1 vector-based plasmids encoding WT DnaA (pKA234) or the DnaA F46A H136A double mutant (pFH) into dnaA46 cells with or without priC. Growth of dnaA46 cells containing pING1 was inhibited at 42℃, but the leaky expression of dnaA from pKA234 enabled the growth of these cells, as previously reported (Sakiyama et al., 2018) (Figure 3). Introduction of pFH moderately supported the cell growth of dnaA46 cells at 42℃. By contrast, dnaA46 ΔpriC cells carrying pFH exhibited little growth at 42℃ (Figure 3).

Role of PriC in growth of DnaA mutant defective in interactions with DnaB. Cell growth of MIT125 (MG1655 dnaA46 tnaA::Tn10) and KRC004 (MG1655 dnaA46 tnaA::Tn10 ΔpriC::frt-kan) cells bearing pING1 plasmid vector (-) or its derivatives pKA234 expressing WT DnaA (WT) or pFH expressing DnaA F46A H136A (F46A H136A). Cells were grown at 30°C overnight and 10-fold-serial dilutions of the cultures (∼109 cells/mL) were incubated on LB agar plates containing ampicillin for 16 h at 30°C and for 14 h at 42°C, respectively. Four independent experiments were performed. +, WT priC; -, ΔpriC::frt-kan.

Western blotting analysis supported that the expression levels of WT DnaA and DnaA F46A H136A proteins in dnaA46 cells bearing pKA234 or pFH were comparable, regardless of the presence or absence of priC+ (Figure 3-figure supplement 1). As DnaA46 is unstable and thermolabile (Shimuta et al., 2004), these cells were grown at 30°C, and were incubated at 42°C for 90 min before Western blotting analysis. The DnaA level in dnaA46 cells bearing the vector pING1 plasmid was minimum. These results suggest that PriC is crucial for the growth of cells defective in specific DnaA-DnaB interactions, possibly because it can bypass the DnaA-DnaB interaction step required for DnaB loading onto oriC ssDNA.

PriC rescues DNA replication inhibition by DiaA overexpression in vivo

DiaA, a homotetrameric protein, binds to multiple DnaA molecules and stimulates DnaA multimer assembly at oriC (Keyamura et al., 2007). The Phe46 residue in DnaA domain I is part of the DiaA-binding site and DnaB binding site. Oversupply of DiaA inhibits timely initiation of replication in cells growing in tryptone medium at 30°C, probably because of binding competition between DiaA and DnaB for DnaA domain I (Ishida et al., 2004; Keyamura et al., 2009). If PriC can bypass the DnaA-DnaB interaction requirement for DnaB loading onto oriC, PriC should suppress inhibition of initiation by DiaA oversupply. To test this possibility, we introduced pBR322 or its derivative encoding diaA (pNA135) into cells with or without priC (Figure 4). priC+ cells carrying pBR322 or pNA135 similarly grew on LB agar plates containing ampicillin between 25°C and 42°C (Figure 4A and Figure 4-figure supplement 1). However, cell growth of ΔpriC cells carrying pNA135, but not pBR322, was moderately inhibited at 25°C (Figure 4A), supporting our conjecture that PriC can bypass the requirement for the DnaA-DnaB interaction.

PriC stimulation of DNA replication initiation inhibited by DiaA oversupply. (A) Cell growth of MG1655, KRC002 (MG1655 ΔpriC::frt-kan), and FF001 (MG1655 Δrep::frt-kan) cells bearing pBR322 or pNA135 (pBR322 derivatives carrying diaA). Overnight cultures (∼109 cells/mL) of MG1655 and KRC002 cells bearing pBR322 or pNA135 were 10-fold serially diluted and incubated on LB agar plates containing ampicillin for 40 h at 25℃. The results of three independent experiments were consistent, and one which is shown. (B) Flow cytometry analyses of MG1655 and KRC002 (MG1655 ΔpriC::frt-kan) cells bearing pBR322 or pNA135. Cells were exponentially grown at 30℃ in LB medium with ampicillin, followed by further incubation with rifampicin and cephalexin to allow run-out of chromosomal DNA replication. DNA contents were quantified by flow cytometry, and cell size (mass) at the time of drug addition was measured by a Coulter counter. Mean mass, ori/mass ratio, doubling time (Td), and asynchrony index (A.I.: the percentage of cell numbers with non-2n copies of oriC per the cell numbers with 2n copies of oriC) of each strain are indicated at the top right of each panel with standard deviations. Three to four independent experiments were performed. +, WT priC; -, ΔpriC::frt-kan. (C) Flow cytometry analyses of MG1655, KRC003 (MG1655 ΔpriC::frt), SA103 (MG1655 ΔdiaA::frt-kan), and KRC006 (MG1655 ΔpriC::frt, ΔdiaA::frt-kan) cells. Cells were grown and analyzed as described above. Three independent experiments were performed. For “priC” column, +, WT priC; -, ΔpriC::frt. For “diaA” column, +, WT diaA; -, ΔdiaA::frt-kan.

Rep helicase stimulates PriC-mediated rescue of arrested replication forks containing nascent strands (Heller & Marians, 2005b, 2005a; Sandler, 2000). To test the contribution of Rep helicase to PriC-promoted rescue of blocked replication initiation, we introduced pBR322 or pNA135 into Δrep cells. Unlike ΔpriC cells, Δrep cells carrying pNA135 grew similarly to Δrep cells carrying pBR322 (Figure 4A), suggesting no involvement of Rep in the PriC-promoted rescue of blocked replication initiation. This is consistent with the features of unwound oriC, which has a fork structure similar to that of the replication fork but without nascent strands, and the specific role of Rep in rescuing abandoned replication forks with nascent strands in the PriC-dependent pathway. Rep is specifically required for the unwinding of the nascent leading strand of abandoned replication forks, which is needed to expand the ssDNA region required for entry of PriC (Heller & Marians, 2005b, 2005a).

We further analyzed replication initiation in ΔpriC cells carrying pNA135 at 30°C by flow cytometry (Figure 4B). Under these conditions, priC+ cells bearing pBR322 mainly had eight or four oriC copies per cell with eight-oriC cells predominating (Figure 4B). The number of oriC copies per cell was similar to that in priC+ cells bearing pNA135, with only slight increase of asynchronous initiation, as shown by the slight increase in the number of cells with non-2n oriC copies, which is basically consistent with our previous data (Ishida et al., 2004). However, in ΔpriC cells bearing pNA135, severe inhibition of initiation occurred, as shown by the substantial decrease in the number of eight-oriC cells and the large increase of the number of cells with only one to three oriC copies (Figure 4B). This result supports the proposed function of PriC at oriC mentioned above.

To test whether PriC rescues replication initiation inhibited by deletion of the diaA gene, we conducted similar flow cytometry analysis of ΔpriC, ΔdiaA, and ΔpriC ΔdiaA cells (Figure 4C). ΔdiaA cells showed severe asynchrony of replication initiation, as shown by the increase in the number of five-to-seven-oriC cells, which supports the stimulatory role of DiaA in DnaA-oriC complex formation (Figure 4C). However, ΔpriC ΔdiaA cells did not show more asynchronization of replication initiation than ΔdiaA cells, as shown by the similar asynchrony index (A.I.). These results suggest that PriC does not support or bypass DnaA-oriC complex formation.

PriC stimulates replication initiation in cells with oriC sequence truncations

Right DnaA-subcomplex plays a stimulatory role in DnaB helicase loading (Sakiyama et al., 2022; Stepankiw et al., 2009). The loading of two DnaB hexamers at oriC ssDNA regions requires multiple steps that depend on the distinct functions of two DnaA subcomplexes (Figure 1): the Left-DnaA subcomplex stably unwinds M-R regions of the DUE and the Right-DnaA subcomplex expands the unwound region to the AT-L region, supporting efficient DnaB helicase loading (Sakiyama et al., 2017, 2022; Yoshida et al., 2023). The expanded ssDNA regions in this open oriC complex efficiently promote the loading of one DnaB hexamer onto the lower (A-rich) strand M-R region and a second DnaB hexamer onto the upper (T-rich) strand of the DUE (Fang et al., 1999; Sakiyama et al., 2022). In the absence of the Right-DnaA subcomplex, the AT region of the AT-L region assists in DnaB loading (Sakiyama et al., 2022)

To determine whether PriC rescues defects in DnaB loading when the oriC sequence is altered to prevent Right-DnaA subcomplex formation, we first assessed the cell growth of cells lacking the Middle- and Right-DOR of oriC (Left-oriC) (Figure 5A). The Left-oriC mutant formed colonies at 37°C but showed severe growth defects at 25°C and 30°C (Figure 5B). At 37°C, thermal energy might be sufficient to expand DUE unwinding into the AT region and allow DnaB loading in the absence of the Right-DnaA subcomplex (Sakiyama et al., 2022), but not at 25°C or 30°C. By contrast, the cell growth of the Left-oriC ΔpriC double mutant was markedly compromised at 37°C and moderately reduced at 25°C and 30°C (Figure 5B). These findings support the idea that PriC functions to rescue defective processes in DnaB loading caused by deletions of the oriC sequence that prevent DnaA subcomplex formation.

Stimulation of replication initiation by PriC in Right-DnaA subcomplex-defective cells (A) Schematic structure of Left-oriC and R4Tma oriC. AT-rich region, DUE, and DOR are indicated by a gray dotted box, a gray box, and an open box, respectively. Filled and open arrowheads in the DOR show DnaA boxes with the full consensus sequence or mismatches. In addition, a striped box indicates the IBR. The regions of each mutant oriC are indicated below the structure as open boxes. The position of the TmaDnaA box substitution is indicated by a yellow arrowhead. (B) Cell growth of WT oriC and Left-oriC cells with or without priC. Overnight cultures (∼109 cells/mL) of NY20-frt (WT oriC; WT) and NY20L-frt (Left-oriC; Left) cells with (+) or without (-) priC were 10-fold serially diluted and incubated on LB agar plates for 24 h at 25°C, for 16 h at 30°C, and for 14 h at 37°C. +, WT priC; -, ΔpriC::frt-kan. Three independent experiments were performed. (C) Cell growth of WT oriC and oriC R4-box mutant cells with or without priC. NY20-frt (WT oriC) and NY24-frt (oriC with R4Tma substitution, R4Tma) cells with (+) or without (-) priC were grown as described in panel A. +, WT priC; -, ΔpriC::frt-kan. Four independent experiments were performed. (D) Flow cytometry analyses of WT and R4Tma cells with or without priC. Cells were exponentially grown at 30℃ in LB medium with ampicillin and analyzed as described in Figure 4. Representative histograms from four independent experiments are shown. Mean mass, ori/mass, and Td of each strain are indicated at the top right of each panel with standard deviations.

Next, similar experiments were performed using a mutant in which the R4 DnaA box within Right-DOR was substituted with a sequence (R4Tma) defective in binding of E. coli DnaA (Figure 5A and 5C) (Noguchi et al., 2015; Sakiyama et al., 2022). We reasoned that formation of the Right-DnaA subcomplex would be inhibited by the introduction of R4Tma and PriC would be required for robust initiation in this context. As the Left-oriC strain, the priC+ R4Tma strain showed normal cell growth. By contrast, cell growth of the R4Tma ΔpriC double mutant was impaired moderately at 25°C and slightly at 30°C (Figure 5C), indicating that PriC can rescue the blocked replication initiation caused by the absence of the R4 box. The reduced cell growth of R4Tma ΔpriC strain was alleviated at higher temperatures, similar to the phenotype of the Left-oriC mutant.

Furthermore, we assessed replication initiation in R4Tma cells at 30°C using flow cytometry (Figure 5D). R4Tma cells grew slightly slower than WT oriC cells and showed clear inhibition of initiation; in contrast to the WT oriC strain, the R4Tma mutant cell population contained more four-oriC cells and fewer eight-oriC cells, and showed severe asynchronous initiation. Notably, these negative effects of the R4Tma mutation were amplified by deletion of priC, i.e., the number of one to three-oriC cells increased and the ori/mass ratio was further reduced.

We next analyzed the role of PriC in cells lacking the AT-L region of oriC (subATL oriC) (Figure 6A). L-DUE and the flanking AT region assist in DnaB helicase loading by stimulating oriC unwinding, which is essential in a strain defective in Right-DnaA subcomplex formation (Sakiyama et al., 2022). SubATL oriC cells and WT oriC cells grew similarly on LB agar plates between 25°C and 37°C, regardless of the presence of ΔpriC (Figure 6B). However, flow cytometry analysis revealed that deletion of priC specifically inhibited initiation in subATL oriC cells (Figure 6C); the number of eight-oriC cells was lower while the number of four-to-seven oriC cells was higher in subATL oriC ΔpriC cells than in priC+ subATL oriC cells. Taken together, these results are consistent with the idea that PriC restores initiation and DnaB loading at such truncated oriC sequences.

Stimulation of replication initiation by PriC in cells with a deletion in the ssDNA-expanding region of oriC (A) Schematic structure of subATL oriC as shown in Figure 5. The oriC region including the subATL oriC is indicated below the structure as a white box. (B) Cell growth of WT oriC and oriC subATL mutant cells with or without WT priC. NY20-frt (WT oriC) and NY20ATL-frt (oriC lacking AT and L sequences, subATL) with (+) or without (-) WT priC were grown as described in Figure 5. +, WT priC; -, ΔpriC::frt-kan. Three independent experiments were performed. (C) Flow cytometry analyses of WT oriC and oriC subATL mutant cells with or without WT priC. Cells were exponentially grown at 30℃ in LB medium and analyzed as described in Figure 4. Representative histograms from three independent experiments are shown. Mean mass, ori/mass, and Td of each strain are indicated at the top right of each panel with standard deviations.

PriC loads DnaB at oriC unwound by the DnaA complex in an in vitro reconstituted system

To analyze the mechanism of PriC rescue of blocked initiation at oriC, we assessed PriC activity in an in vitro reconstituted system for DnaB loading and DnaB-promoted DNA unwinding using the supercoiled circular form (form I) of the oriC plasmid pBSoriC and purified proteins, namely DnaA, N-terminally histidine-tagged DnaB (His-DnaB), DnaC, IHF, SSB, and gyrase. In this system, DnaA and IHF unwind the DUE and stimulate His-DnaB loading to the ssDUE region with the aid of DnaC. The loaded His-DnaB expands the ssDNA regions in concert with the ssDNA-binding activities of SSB. Concomitantly, the positive supercoiling generated through DNA unwinding by His-DnaB is resolved by DNA gyrase, resulting in the formation of plasmid DNA topoisomers (form I*). Form I* can be separated from form I by agarose gel electrophoresis (Baker et al., 1986; Sakiyama et al., 2022).

First, we assessed the effect of PriC on WT DnaA. WT DnaA stimulated form I* formation in the absence of PriC (Figure 7A, lanes 1 and 5) and increasing concentrations of PriC moderately inhibited WT DnaA-mediated form I* formation (Figure 7A, lanes 5 to 8), probably because of competition between DnaA and PriC for binding to DnaB. Next, we focused on the DnaA F46A H136A double mutant. The DnaA F46A and DnaA H136A mutant proteins alone are reported not to support form I* formation (Keyamura et al., 2009; Sakiyama et al., 2018). Consistently, the DnaA F46A H136A double mutant protein was inactive in form I* formation (Figure 7A, lane 9). However, when PriC was added to the assay, DnaA F46A H136A promoted form I* formation to a level comparable to that of WT DnaA in the presence of PriC (Figure 7A, lanes 5-12 and 7B), indicating that PriC rescues defective DnaB loading by DnaA F46A H136A-oriC complexes. These findings are consistent with the idea that PriC can bypass the strict reliance on DnaA-DnaB interactions for DnaB loading at oriC.

DnaB loading onto the ssDUE by PriC. (A and B) In vitro reconstituted system for DnaB loading. Form I of oriC plasmid (pBSoriC; 1.6 nM) was incubated for 15 min at 30℃ with the indicated amount of PriC in the presence of ATP-DnaA or its mutant derivatives, DnaB, DnaC, IHF, gyrase, and SSB. The resulting plasmids were purified and analyzed using agarose gel electrophoresis. A representative image of three independent experiments is shown in the black/white-inverted mode (A). Band intensities of each lane in the gel image were analyzed by densitometric scanning. The percentages of form I* oriC plasmid per input DNA are shown as ‘Form I* (%)’. Mean and standard deviations are shown (B). Abbreviations in panel B: -, no addition of DnaA; WT, wild-type DnaA; FH, DnaA F46A H136A; VA, DnaA V211A. (C and D) In vitro reconstituted system containing DiaA. Form I of pBSoriC plasmid was used as described above except for the addition of the indicated amounts of DiaA. A representative image of three independent experiments is shown in black/white-inverted mode (C). Band intensities of each lane in the gel image were analyzed by densitometric scanning. The percentages of form I*oriC plasmid together with the mean and standard deviations are shown as described above (D).

To corroborate this idea, we also determined whether PriC rescues blocked initiation resulting from unstable DUE unwinding. oriC complexes containing DnaA V211A cannot bind to the ssDUE and consequently fails in stable DUE unwinding (Ozaki et al., 2008). The DnaA V211A mutant was virtually inactive for form I * formation irrespective of the presence or absence of PriC (Figure 7A, lanes 13-16, and 7B), indicating that PriC cannot efficiently rescue a defect of DUE unwinding.

We further evaluated the DnaB loading activity of PriC in the presence of DiaA using a similar in vitro reconstituted system. Previously, DiaA was shown to stimulate oriC DUE unwinding, but to inhibit form I* formation, because DiaA-DnaA binding competitively inhibits DnaB-DnaA binding; DiaA and DnaB share the same binding site in DnaA domain I (Figures 7C lanes 1, 3, 5, and 7 and 7D) (Keyamura et al., 2009). PriC moderately stimulated form I* formation specifically in the presence of both DnaA and DiaA (Figures 7C lanes 7 and 8, and 7D). These results further support the idea that PriC can bypass the specific requirement for the DnaA-DnaB interaction in DnaB loading at oriC. DiaA bound to oriC-DnaA complexes might be a physical obstacle reducing efficiency of DnaB loading by PriC.

In addition, we analyzed the DUE unwinding activities of DnaA mutants using the in vitro reconstituted system. As previously reported, DnaA V211A is largely inactive in DUE unwinding, but DnaA F46A H136A unwinds the DUE at a level comparable to that of WT DnaA (Figure 7-figure supplement 2A and 2B). PriC did not stimulate DUE unwinding irrespective of the presence of WT DnaA (Figure 7-figure supplement 2C and 2D). These results demonstrate the functional specificity of PriC in rescuing failed DnaB loading at oriC ssDNA and are consistent with the PriC mechanism mentioned above.

Role for PriC in cSDR

Finally, to analyze the role of PriC in other types of replication initiation, we examine whether PriC contributes to the initiation of cSDR, which does not require dnaA or oriC and is activated in cells lacking rnhA or recG. Although the PriA-PriB-DnaT primosome complex has previously been reported to be involved in the initiation of cSDR (Heller & Marians, 2006), it is unclear whether PriC has any role in this type of initiation.

Unlike dnaA46 cells, dnaA46 rnhA::cat double mutant cells grew even at 40°C (Figure 8A), indicating that these cells were engaged in cSDR, as previously reported (Hinds & Sandler, 2004). The cell growth of ΔpriC dnaA46 rnhA::cat triple mutant was severely inhibited at 40°C (Figure 8A), suggesting that PriC contributes to the growth of dnaA46 rnhA::cat mutant cells.

PriC stimulation of cSDR-dependent cell growth. (A) Cell growth abilities. Cells of MIT125 (MG1655 dnaA46), MIT125c (MG1655 dnaA46 rnhA::cat), KRC004 (MG1655 dnaA46 ΔpriC::frt-kan), and KRC004c (MG1655 dnaA46 ΔpriC::frt-kan rnhA::cat) were grown at 30℃ overnight and 10-fold-serial dilutions of the overnight cultures (∼109 cells/mL) were incubated for 16 h at 30℃ and for 36 h at 40℃ on LB agar plates. Three independent experiments were performed. +, WT; -, deletion. (B) oriC and ter copy numbers. MIT125, MIT125c, KRC004, and KRC004c cells growing exponentially at 30℃ were further incubated for 90 min at 40℃ in LB medium. Samples were withdrawn before and after the 40℃ incubation. The genome DNA of each sample was extracted by boiling the cells for 5 min at 95℃. The relative copy numbers of oriC (84.6 min) and ter (32.4 min) to that of ypaB (50.5 min) were quantified using real-time qPCR. The data and averages of three independent experiments are shown with standard deviations.

To determine whether PriC contributes to the initiation of dnaA46-oriC replication or cSDR, we calculated the relative copy numbers of oriC (84.6 min) and ter (32.4 min: corresponding to insQ) using real-time quantitative PCR (qPCR). The ypaB (50.5 min) located at the middle position between oriC and ter was used as a reference. To calculate the copy number of the chromosomal ter region, the preferential initiation site of cSDR (Brochu et al., 2018; Maduike et al., 2014), dnaA46 rnhA::cat cells growing exponentially at 30°C were shifted to 40°C for 90 min and the copy number before and after incubation at 40°C was calculated. Consistent with the cell growth data (Figure 8A), the ter copy number ratio of the dnaA46 rnhA::cat double mutant increased after 40°C incubation, whereas the copy number ratio of the dnaA46 mutant expressing WT rnhA remained the same (Figure 8B). The ter copy number ratio of the dnaA46 rnhA::cat ΔpriC triple mutant strain was also increased, indicating that cSDR could occur even in the absence of PriC.

To monitor initiation of DnaA-oriC replication, we determined the oriC copy number ratio (Figure 8B). At the permissive temperature, despite having similar growth rates, the oriC copy number ratio of the dnaA46 ΔpriC double mutant cells was lower than that of the dnaA46 single mutant, confirming the importance of PriC for replication initiation in dnaA46 cells, as shown in Figure 2. In the priC+dnaA46 rnhA::cat mutant cells, the oriC copy number ratio was reduced, which is explained by activation of cSDR and increase of the relative copy number of the reference ypaB locus (50.5 min chromosomal map position). Together, these results suggest the idea that PriC stimulates replication fork progression during cSDR.

To consolidate these findings, we performed the chromosome loci copy-number analysis using whole-genome sequencing. The dnaA46 cells and its derivative cells were grown exponentially at 30°C, followed by incubation at 40°C for 90 min prior to DNA extraction. In growing WT cells, the highest copy numbers are observed near the oriC-proximal region and the lowest are near the ter-proximal region, resulting in a coverage bias from oriC to ter (Brochu et al., 2018; Maduike et al., 2014).

However, in dnaA46 cells incubated at 40°C, the coverage bias was largely diminished due to the run-out replication (Figure 8-figure supplement 1A). Notably, dnaA46 rnhA::cat cells incubated at 40°C exhibited discreate coverage peaks at the preferential initiation sites of cSDR near the ter (or insQ) region with a decreasing coverage bias to oriC. Similar profiles were observed in the ΔpriC mutant background, supporting the idea that cSDR can occur even in the absence of PriC (Figure 8-figure supplement 1A). Moreover, in the dnaA46 rnhA::cat background, the absence of priC appeared to decrease the coverage level at genomic positions near the 1.3 MB and 4 MB map-positions, which are slightly outside of ter and oriC, respectively (Figure 8-fugure supplement 1B). This observation may reflect the role for priC in the replication fork progression.

Discussion

In growing cells, replication initiation can be blocked under challenging conditions. However, the mechanisms that rescue blocked initiations are largely unknown. To elucidate such mechanisms, we investigated PriC-promoted replication initiation under various challenging conditions. We showed that PriC was necessary for the optimal growth of dnaC2 cells at the semi-restrictive temperature and for replication initiation even at the permissive temperature (Figure 2), suggesting that PriC contributes to replication initiation rescue in dnaC2 cells by facilitating DnaB helicase loading at oriC. Furthermore, we observed that PriC was necessary for the optimal growth of cells in which the DnaA-DnaB interaction was inhibited by DiaA overexpression or DnaA F46A H136A double mutations (Figures 3 and 4). In addition, PriC was stimulatory for the cell growth and replication initiation bearing oriC mutations that impair efficient DnaB loading (Figures 5 and 6). These results, together with results of form I* formation assays (Figures 7), suggest that PriC bypasses the need for the DnaA-DnaB interaction by binding to unwound oriC DNA and recruiting DnaB to facilitate its loading onto ssDNA (Figure 9). This mechanism is similar to that proposed for the rescue of abandoned replication forks by PriC (Figure 1E).

Model for PriC-promoted replication initiation.

PriC-promoted rescue of blocked DnaA-oriC dependent replication initiation. oriC, DnaA, IHF, DnaB, and DnaC are illustrated as depicted in Figure 1C. When tight DnaA binding of DiaA (dark gray circle) (or DnaA F46A H136A double mutations) inhibits DnaB recruitment, PriC (Brilliant red circle) binds to a stably unwound DUE strand, recruiting a DnaB-DnaC complex to lower strand of DUE instead of Left-DnaA subcomplex. When a sufficient ssDNA region is exposed, PriC then recruits the DnaB-DnaC complex to the opposite strand.

Also, PriC contributes to replication initiation rescue in dnaA46 cells of 37°C. It is presumable that DnaA46 protein becomes partially denatured at the sub-permissive temperature of 37°C. This partial denaturation should impair both origin unwinding and helicase loading, though not to the extent that cell viability is lost. The priC deletion should further exacerbate helicase loading defects by inhibiting the bypass mechanism (Figure 9), resulting in the lethality of dnaA46 cells at this temperature.

There might remain an alternative possibility that PriC could stabilize the DnaA-DnaB interaction for recusing the impaired DnaB loading process. However, we believe that this possibility is not very likely. Given that interactions between DnaA and DnaB during DnaB loading to oriC are highly dynamic and involve regulated multiple steps, the stabilization of the DnaA-DnaB interaction by PriC, even if it occurs, carries a considerable risk of inhibiting the DnaB loading by constructing abortive complexes. In addition, DnaA-DiaA binding is very tight and stable (Keyamura et al.,2007, 2009). Even if WT DnaA and WT DnaB are present, PriC can rescue the initiation defects of oriC mutants. Based on these facts and the known characteristics of PriC, it is more reasonable to infer that PriC provides a bypass of DnaB loading at oriC, similar to the mechanism at the stalled replication fork (Figures 1E and 9).

The finding that PriC failed to stimulate form I* formation in the presence of DnaA V211A, which lacks DUE unwinding activity (Figures 7), provides mechanistic insight into PriC-dependent replication initiation at oriC. We propose that PriC-dependent replication initiation possibly begins with the recognition of a stably unwound ssDNA region by PriC (Figure 9). Consistent with this perspective, PriC ssDNA-binding activity and ssDNA regions have been reported to be important for PriC-promoted replication fork rescue (Heller & Marians, 2006; McMillan & Keck, 2024). The failure of PriC to stimulate replication initiation in ΔdiaA cells (Figure 4C) is also potentially explained by the reduced DUE unwinding caused by destabilization of the DnaA-oriC complex in ΔdiaA cells. Based on this, we suggest that the stably unwound ssDNA region in oriC serves as a distinctive feature for PriC-promoting DnaB loading at oriC for replication initiation rescue, and distinguishes it from PriA-dependent replication restart, which requires the specific structure of the replication fork DNA consisting of sister dsDNA strands (Heller & Marians, 2005a; Windgassen et al., 2018). Unlike PriA, PriC can interact with the simple ssDNA to reload DnaB (Figure 1E).

In this study, we found that cells with DnaA F46A H136A double mutations grew well, but only when the cells expressed PriC. As for the DnaA F46A single mutant, our previous studies show that DnaA F46A has a limited residual activity in vivo (unlike in vitro), and allows slow growth of cells (Abe et al., 2007). The growth of cells expressing the DnaA H136A single mutant was severely inhibited even when the cells expressed PriC (Sakiyama et al., 2018). The His136 residue is located within the weak, secondary DnaB interaction region in DnaA, and is crucial for DnaB loading onto oriC ssDNA. Although domain I in DnaA H136A can stably tether DnaB-DnaC complexes to DnaA complexes on oriC (Sakiyama et al., 2018), the complexes fail to load DnaB onto oriC ssDNA even in the presence of PriC. It is possible that the interaction between PriC and DnaB is inhibited by stable DnaB binding to DnaA domain I. This idea is consistent with the in vivo feature of DnaA F46A single mutant.

Conversely, the PriC-DnaB interaction may inhibit the interaction between DnaA domain I and DnaB, as suggested by the inhibitory effect of PriC on DnaB loading in vitro in the presence of WT DnaA (Figure 7). However, this inhibitory effect of PriC was not observed in vivo, suggesting that PriC-DnaB binding could involve still unknown factors and regulatory mechanisms.

PriC supported cSDR-dependent growth in dnaA46 rnhA::cat double mutant cells without stimulating the initiation of cSDR (Figure 8). In this mutant, cSDR predominantly initiates from the ter region, resulting in head-on collisions between the replisome and transcription complexes in rrn operons (Maduike et al., 2014). This suggests that in cSDR cells PriC specifically rescues abandoned forks including those generated during head-on conflicts between the replisome and transcription complexes.

Bacillus subtilis (B. subtilis), which expresses a PriA homolog but not a PriC homolog, employs a PriA-independent fork rescue pathway (Bruand et al., 2001). In this pathway, B. subtilis DnaC (BsuDnaC) helicase is reloaded using the helicase loader DnaI together with B. subtilis DnaB (BsuDnaB) and DnaD coloaders. The ssDNA-binding activity of BsuDnaB has a crucial role in this PriA-independent pathway (Bruand et al., 2005). SsDNA-bound BsuDnaB remodels the SSB-ssDNA complex with DnaD and recruits BsuDnaC. These functions of BsuDnaB are similar to the proposed functions of PriC in replication fork rescue in E. coli. In addition, BsuDnaC helicase is recruited to oriC by concerted actions of B. subtilis DnaA and the DnaD-BsuDnaB complex, which has an essential role in chromosome replication initiation (Jameson & Wilkinson, 2017), suggesting a similar role of BsuDnaB to PriC in E. coli cells expressing DnaA F46A H136A. These similarities between PriC and BsuDnaB in these two evolutionarily distant bacterial species suggest that the mechanism of PriC-promoted helicase loading is conserved among bacterial species despite the absence of sequence conservation of PriC and BsuDnaB homologs.

Based on our results, we propose the abnormal competition between DnaB and DiaA for DnaA domain I could represent a form of intrinsic replication initiation stress in bacteria with conserved DiaA homologs. This type of stress could also occur in ε-proteobacterial species, such as Helicobacter pylori, because they express HobA, a DiaA-functional homolog (Natrajan et al., 2007; Zawilak-Pawlik et al., 2011). Similarly, YfdR, encoded by E. coli prophages, inhibits replication initiation by competing with DnaB for binding to DnaA domain I (Noguchi & Katayama, 2016), suggesting that such extraneously introduced inhibitors could trigger replication initiation stress.

Even in human, ORC1, ORC2, and ORC5, the essential components of the eukaryotic replication initiation complex, are not essential in some cancer cell-lines (Shibata & Dutta, 2020; Struhl et al., 2016), suggesting that mechanisms of replication initiation rescue may also operate beyond the bacterial kingdom. Therefore, further investigation of the initiation rescue processes and the factors involved in diverse organisms from bacteria to human will be important for a full understanding of the common principles and diverse mechanisms that ensure robust initiation of chromosomal DNA replication.

Materials and Methods

Plasmids, proteins, and strains

The plasmids used in this study are listed in Key Resource Table. pKA234, pKW44-1, pNA135, pBSoriC, pET22b(+)-priC, and pTKM601 were described previously (Aramaki et al., 2015; Ishida et al., 2004; Kawakami et al., 2005; Keyamura et al., 2009; Kubota et al., 1997; Ozaki et al., 2008). For the construction of pFH, an alanine substitution was introduced into pH136A with specific mutagenic primer sets using the QuikChange site-directed mutagenesis protocol (Stratagene [Agilent], Agilent, La Jolla, CA, United States) as described previously (Keyamura et al., 2007; Sakiyama et al., 2018).

WT DnaA and its derivative proteins were overproduced in E. coli strain KA450 from pKA234, pKW44-1, or pFH and purified as described previously (Noguchi et al., 2015; Ozaki et al., 2008; Sakiyama et al., 2017). PriC protein was prepared as previously reported with minor modifications (Aramaki et al., 2013).

Briefly, PriC protein was overproduced in E. coli strain BL21-codonPlus-RIL by inducing its expression from pET22b(+)-priC using 0.2 mM isopropyl-β-D-1-thiogalactopyranoside (IPTG). The resulting cells were suspended in chilled buffer A (50 mM HEPES-NaOH [pH7.0], 10% sucrose, 1 mM EDTA, 2 mM dithiothreitol (DTT), and 1 mM PMSF) and disrupted by sonication. Proteins in the soluble fraction were precipitated with 0.24 g/mL ammonium sulfate, resuspended in a separation buffer (50 mM imidazole [pH 7.0], 20% glycerol, 2 mM DTT, 1 mM EDTA, and 40 mM ammonium sulfate) and loaded onto a 1 mL HiTrap® SP HP column (Cytiva, Uppsala, Sweden). Bound proteins were eluted with a linear gradient from 0 to 1 M sodium chloride.

All E. coli strains used in this study are listed in Key Resource Table. KYA018, MIT125, MIT162, SA103, NY20 and NY21 were described previously (Kasho & Katayama, 2013; Noguchi et al., 2015; Noguchi & Katayama, 2016). Strains bearing mutant oriC (NY20L and NY20ATL) were constructed using the λRed site-directed recombination system as previously described (Noguchi et al., 2015; Sakiyama et al., 2017, 2022). NY20-frt strain and oriC mutant strains (NY21-frt, NY20L-frt, and NY20ATL-frt) were constructed by eliminating kan from the NY20 strain and strains NY21, NY20L, and NY20ATL. To remove the kanamycin-resistant cassette (kan) from oriC, FLP recombinase encoded on pCP20 was used. Elimination of kan was verified by checking sensitivity to 50 µg/ml kanamycin in LB agar plates. ΔpriC::frt-kan from JW0456-KC and Δrep::frt-kan from JW5604-KC were introduced into MG1655 by P1 transduction, resulting in the construction of KRC002 and FF001, respectively. For construction of other ΔpriC::frt-kan strains (KRC004, KRC005, NY20-priC, NY21-priC, NY20L-priC, NY20ATL-priC), P1 phage lysates prepared from KRC002 were used for transduction of strains MIT125, KYA018, NY20-frt, SYM21-frt, NY20L-frt, and NY20ATL-frt. Transductants were screened on LB agar plates containing 50 µg/mL kanamycin. ΔdiaA::frt-kan or rnhA::cat derivatives (KRC006, MIT125c, and KRC004c) were constructed using a similar protocol except that P1 phage lysates from SA103 (ΔdiaA::frt-kan) or MIT162 (rnhA::cat) were used. KRC003 was generated by removal of kan from KRC002.

Buffers

Buffer P contained 60 mM HEPES–KOH (pH 7.6), 0.1 mM zinc acetate, 8 mM magnesium acetate, 30% [v/v] glycerol, and 0.32 mg/mL bovine serum albumin (BSA). Buffer N contained 50 mM HEPES–KOH (pH 7.6), 2.5 mM magnesium acetate, 0.3 mM EDTA, 7 mM DTT, 0.007% [v/v] Triton X-100, and 20% [v/v] glycerol. Form I* buffer contained 20 mM Tris-HCl (pH 7.5), 125 mM potassium glutamate, 10 mM magnesium acetate, 8 mM DTT, and 0.5 mg/mL BSA.

DUE unwinding assay

DUE unwinding assays were performed essentially as described with minor modifications (Sakiyama et al., 2022; Yoshida et al., 2023). Briefly, DnaA was incubated with ATP in buffer N on ice for 3 min to generate ATP-DnaA. The indicated amount of PriC and ATP-DnaA or its mutant derivatives was incubated for 9 min at 30℃ in 10 µL buffer P containing 5 mM ATP, 125 mM potassium glutamate, 1.6 nM pBSoriC, and 42 nM IHF, followed by further incubation with 1.5 units of P1 nuclease (Wako) for 5 min. The reaction was stopped by the addition of 1% SDS and 25 mM EDTA, and DNA was purified by phenol-chloroform extraction and ethanol precipitation. One-third of each purified DNA was digested with AlwNI (NEB), which yielded 2.6 kb and 1.0 kb fragments after DUE unwinding of pBSoriC. The resultant DNA fragments were analyzed by 1% agarose gel electrophoresis with 1xTris-acetate-EDTA buffer for 30 min at 100 V, followed by ethidium bromide staining. Gel images were taken using the GelDoc GO imaging system (Bio-Rad Laboratories, Hercules, CA), and products derived from unwound plasmids were quantified using ImageJ software.

Form I* assay

This assay was performed as previously described with minor modifications (Baker et al., 1986; Noguchi et al., 2015; Sakiyama et al., 2022). The indicated amount of ATP-DnaA was incubated for 15 min at 30℃ in 12.5 µL of Form I* buffer containing 3 mM ATP, 1.6 nM pBSoriC, 42 nM IHF, 400 nM His-DnaB, 400 nM DnaC, 76 nM GyrA, 100 nM His-GyrB, and 760 nM SSB. The reaction was stopped by the addition of 0.5% SDS, and DNA was purified by phenol-chloroform extraction. Samples were analyzed by 0.65% agarose gel electrophoresis with 0.5xTris-borate-EDTA buffer for 15 h at 23 V, followed by ethidium bromide staining.

Flow cytometry analysis

Flow cytometry analysis was performed essentially as described (Kasho et al., 2014). Briefly, cells were grown at 30℃ in LB medium until the absorbance of the culture (A600) reached 0.1. Portions of the cultures were diluted a thousand-fold into 5 ml LB medium and incubated at 30℃ until the absorbance of the culture (A600) reached 0.1. The remaining portions were further incubated to determine the doubling time (Td) by measuring A600 every 20 min. At A600 0.1, aliquots of the cultures were fixed in 70% ethanol to analyze cell mass using the Multisizer 3 Coulter counter (Beckman Coulter, Brea, CA). The remaining cultures were further incubated for 4 h with 0.3 mg/mL rifampicin and 0.01 mg/mL cephalexin to allow run-out replication of chromosomal DNA. The resultant cells were fixed in 70% ethanol. After DNA staining with SYTOX Green (Thermo Fisher Scientific, Waltham, MA), cellular DNA contents were analyzed on a FACS Calibur flow cytometer (BD Bioscience, Franklin Lakes, NJ). The number of the origins/cell (ori/cell) was determined from the histograms of flow cytometry analysis. The number of ori/cell was divided by the mean cell mass determined by cell mass analysis, resulting in ori/mass ratios. The numbers of cells containing non-2n copy number of oriC were divided by the number of cells containing 2n copy number of oriC to calculate the asynchrony index (A.I.). For the analysis of cells bearing plasmids, LB medium was supplemented with 100 µg/mL ampicillin.

Replication initiation frequency test using synchronized cultures

MIT125 (dnaA46 tnaA::Tn10) cells or its ΔpriC derivative KRC004 were grown at 30℃, a permissive temperature, in LB medium until the absorbance of the culture (A600) reached 0.04, followed by further incubation for 90 min at 42℃, a restrictive temperature. Aliquots of the cultures were fixed in 70% ethanol as “Synchronization” samples. The remaining cultures were further incubated for 5 min to initiate DNA replication, followed by further incubation for 4 h at 42℃ with 0.3 mg/mL rifampicin and 0.01 mg/mL cephalexin to allow run-out replication of chromosomal DNA. The resultant cells were fixed in 70% ethanol as “5 min release” samples. After DNA staining with SYTOX Green (Life Technologies), cellular DNA contents were analyzed on a FACS Calibur flow cytometer (BD Bioscience). Ratios of the origins/cell (ori/cell) were determined from the histograms of flow cytometry analysis.

For synchronization of dnaC2 mutant cells (KYA018 and KRC005), cells were grown in LB medium at 30℃ until the A600 reached 0.04, followed by further incubation for 80 min at 37℃, at the restrictive temperature. The resulting synchronized cultures were released for 5 min at 30℃ and incubated in the presence of 0.3 mg/mL rifampicin and 0.01 mg/mL cephalexin at 30℃ to allow run-out of chromosomal DNA replication.

qPCR for analysis of cSDR

Cells of MIT125 (dnaA46 tnaA::Tn10) cells and MIT125c (dnaA46 tnaA::Tn10 rnhA::cat) or its ΔpriC derivative KRC004 (dnaA46 tnaA::Tn10 ΔpriC::frt-kan) and KRC004c (dnaA46 tnaA::Tn10 rnhA::cat ΔpriC::frt-kan) were grown at 30℃ in LB medium until the absorbance of the culture (A600) reached 0.04 and aliquots of the cultures were withdrawn. The remaining samples were further incubated for 90 min at 40℃ and aliquots of the cultures were withdrawn. These samples were boiled for 5 min at 95℃ and the genome DNA was extracted. The levels of oriC (84.6 min), ter (32.4 min), and ypaB (50.5 min) were quantified by real-time qPCR using TB Green Premix Ex TaqII (Tli RNaseH Plus) (TaKaRa, Shiga, Japan) and the following primers: ORI_1 and KWoriCRev for oriC, qoriK fw and qoriK rev for ter, STM419 and STM420 for ypaB.

Chromosome loci copy-number analysis by whole-genome sequencing

The analysis was performed based on a previous paper with modifications (Brochu et al., 2018). Cells of the dnaA46 mutant derivatives MIT125 (dnaA46 tnaA::Tn10), MIT125c (dnaA46 tnaA::Tn10 rnhA::cat), KRC004 (dnaA46 tnaA::Tn10 ΔpriC::frt-kan) and KRC004c (dnaA46 tnaA::Tn10 rnhA::cat ΔpriC::frt-kan) were grown exponentially at 30℃ in LB medium (50 mL). When the absorbance of the culture (A600) reached 0.04, cells were incubated for 90 min at 40℃ and collected by centrifugation. Cell pellets were dissolved by incubation at 37℃ for 1h in TE buffer (0.6 mL) containing 0.5% SDS and 25 µg/ml proteinase K, and were mixed with 0.2 mL of buffer containing 4% cetyltrimethylammonium bromide, followed by further incubated for 10 minutes at 65℃, phenol-chloroform extraction and isopropanol precipitation. Precipitates were resuspended in Tris (pH7.5) buffer containing 0.1 mg/ml RNaseA, followed by incubation for 1 hour at 37℃ and DNA purification by FastGene Gel/PCR Extraction Kit (Nippon Genetics, Tokyo, Japan). The quality control of genomic DNA, library preparation, and whole-genome sequencing were performed by Rhelixa, Inc. using NEBNext Ultra II DNA Library Prep Kit and Illumina NovaSeq X Plus. The number of reads (150 bp x 2 pair-ended) was at least 6.7 M (3.3 M pairs) per sample. The sequence data were analyzed using Galaxy (https://usegalaxy.org) and the coverage plot was generated essentially as described (Ivanova et al., 2015). Briefly, paired-end reads were mapped using the MG1655 reference genome (NC_000913) and the BWA-MEM algorithm, followed by calculation of the number of reads per genome coverage. DNA-sequence data are deposited in the NCBI Sequence Read Archive (Accession number: PRJNA1222470).

Immunoblot analysis

The assay was performed as described previously with minor modifications (Kawakami et al., 2005; Sakiyama et al., 2017; Yoshida et al., 2023). Briefly, MIT125 (dnaA46 tnaA::Tn10) and KRC004 (dnaA46 tnaA::Tn10 ΔpriC::frt-kan) cells bearing pING1, pKA234, or pFH were grown at 30°C in LB medium including 100 µg/ml ampicillin until the absorbance of the culture (A600) reached 0.05, followed by incubation at 42°C for 90 min. Cells were harvested by centrifugation and dissolved in SDS sample buffer to adjust the calculated A600 of 0.5. Proteins were separated using SDS-12% polyacrylamide gel electrophoresis and were transferred to polyvinylidene difluoride membranes (Millipore) using a semi-dry blotting method (Bio-Rad). After blocking at 30°C for 30 min in 3% gelatin-TBS (3% gelatin, 20 mM Tris–HCl, 500 mM NaCl, pH 7.5), the membrane was incubated overnight at 4°C in 1% gelatin-TTBS (20 mM Tris–HCl, 500 mM NaCl, 0.05% Tween-20, pH 7.5) containing anti-DnaA rabbit antiserum (1:3000). After washing, the membrane was incubated at 30°C for 1 h in 1% gelatin-TTBS containing goat anti-rabbit IgG antibody conjugated to alkaline phosphatase (Bio-Rad), followed by development using AP Conjugate substrate Kit (Bio-Rad).